---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# goldilocks

<!-- badges: start -->
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/graemeleehickey/goldilocks?branch=master&svg=true)](https://ci.appveyor.com/project/graemeleehickey/goldilocks)
[![CRAN status](https://www.r-pkg.org/badges/version/goldilocks)](https://CRAN.R-project.org/package=goldilocks)
[![R build status](https://github.com/graemeleehickey/goldilocks/workflows/check-standard/badge.svg)](https://github.com/graemeleehickey/goldilocks/actions)
[![R build status](https://github.com/graemeleehickey/goldilocks/workflows/R-CMD-check/badge.svg)](https://github.com/graemeleehickey/goldilocks/actions)
<!-- badges: end -->

The goal of `goldilocks` is to implement the Golilocks Bayesian adaptive design
proposed by Broglio et al. (2014) for time-to-event endpoint trials, both one-
and two-arm, with an underlying piecewise exponential hazard model.

The method can be used for a confirmatory trial to select a trial's sample size
based on accumulating data. During accrual, frequent sample size selection
analyses are made and predictive probabilities are used to determine whether the
current sample size is sufficient or whether continuing accrual would be futile.
The algorithm explicitly accounts for complete follow-up of all patients before
the primary analysis is conducted.

Broglio et al. (2014) re refer to this as a *Goldilocks trial design*, as it is
constantly asking the question, "Is the sample size too big, too small, or just
right?"

## References

Broglio KR, Connor JT, Berry SM. Not too big, not too small: a Goldilocks
approach to sample size selection. *Journal of Biopharmaceutical Statistics*,
2014; **24(3)**: 685–705.

## Installation

You can install the development version of `goldilocks`
[GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("graemeleehickey/goldilocks")
```
