---
title: "ADVENT: a published Goldilocks design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ADVENT: a published Goldilocks design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "advent-cache/",
  fig.width = 7,
  fig.height = 4
)
set.seed(46012244)
```

```{r setup, message=FALSE}
library(goldilocks)
```

The ADVENT trial is a useful published example of a Goldilocks adaptive sample size design. ADVENT compared pulsed field ablation (PFA) with conventional thermal ablation for patients with drug-resistant paroxysmal atrial fibrillation. It was registered as [NCT04612244](https://clinicaltrials.gov/study/NCT04612244), its design was published in *Heart Rhythm O2* (Reddy et al., 2023), and its primary results were published in *The New England Journal of Medicine* (Reddy et al., 2023).

The design paper states that the sample size was determined adaptively using a Goldilocks design, citing Broglio et al. (2014). The [ADVENT statistical analysis plan (SAP)](https://cdn.clinicaltrials.gov/large-docs/44/NCT04612244/SAP_001.pdf) clarifies that the possible sizes of the modified intent-to-treat (mITT) effectiveness population were 350, 450, 550, 650, and 750. At each enrollment milestone, the trial calculated the predictive probability that the trial would eventually show noninferiority for both primary endpoints. The trial could then stop enrollment for predicted success, stop for futility, or continue to the next milestone.

This vignette uses ADVENT as a worked example for `goldilocks`. It focuses on how the published design maps to package arguments. The goal is not to recreate the sponsor's full statistical analysis plan exactly. Some details, including site-stratified randomization, analysis-population exclusions, subject-linked missingness, and the joint co-primary endpoint stopping rule, are simplified here.

## Trial overview

ADVENT was a multicenter, prospective, single-blind, randomized controlled noninferiority trial. Randomized subjects were assigned 1:1 to PFA or standard-of-care thermal ablation, where the thermal arm used either radiofrequency ablation or cryoballoon ablation depending on site. The first 1 to 3 subjects at each site were nonrandomized roll-in subjects and are not part of the randomized comparison modeled below.

```{r trial-summary, echo=FALSE}
trial_summary <- data.frame(
  Feature = c(
    "Population",
    "Treatment arm",
    "Control arm",
    "Randomization",
    "Follow-up",
    "Adaptive sample sizes",
    "Primary effectiveness endpoint",
    "Primary safety endpoint"
  ),
  `Reported ADVENT design` = c(
    "Drug-resistant paroxysmal atrial fibrillation",
    "Pulsed field ablation",
    "Thermal ablation by radiofrequency or cryoballoon ablation",
    "1:1 after nonrandomized roll-in subjects",
    "12 months",
    "350, 450, 550, 650, or 750 mITT subjects",
    paste(
      "Treatment success: acute procedural success and freedom from",
      "specified chronic failures through 12 months"
    ),
    paste(
      "Composite device- or procedure-related serious adverse events,",
      "including selected acute and chronic events"
    )
  ),
  check.names = FALSE
)

knitr::kable(trial_summary)
```

The final published randomized cohort included 305 subjects assigned to PFA and 302 assigned to thermal ablation. In the primary results paper, PFA was noninferior to thermal ablation for both primary effectiveness and primary safety.

The published design also included a trial-flow figure. The following diagram recreates the parts of that flow that matter for the package mapping.

```{r trial-flowchart, echo=FALSE, message=FALSE, out.width='100%'}
DiagrammeR::grViz("
digraph advent_trial_flow {
  graph [rankdir = TB, bgcolor = transparent, ranksep = 0.55, nodesep = 0.35]
  node  [shape = box, style = rounded, fontname = Helvetica, fontsize = 14,
         margin = 0.12]
  edge  [fontname = Helvetica, fontsize = 12]

  screen [label = 'Eligible patients\\nwith drug-resistant PAF']
  rollin [label = 'Site roll-in\\n1 to 3 PFA subjects']
  random [label = 'Randomized comparison\\n1:1 allocation']
  pfa    [label = 'PFA arm']
  thermal [label = 'Thermal ablation arm\\nRF or cryoballoon']
  blank  [label = '90-day blanking period']
  follow [label = 'Follow to 12 months']
  endpoints [label = 'Co-primary endpoints\\neffectiveness + safety']

  screen -> rollin
  rollin -> random
  random -> pfa
  random -> thermal
  pfa -> blank
  thermal -> blank
  blank -> follow
  follow -> endpoints
}
", width = "100%", height = "620px")
```

In the code below, roll-in subjects and randomized subjects excluded from the analysis population are ignored. The simulation starts directly with the modeled 1:1 analysis population.

## The Goldilocks flow

The ADVENT design paper includes a flowchart for the adaptive sample size algorithm. The chart below recreates the logic in package terms rather than copying the published image.

```{r adaptive-flowchart, echo=FALSE, message=FALSE, out.width='100%'}
DiagrammeR::grViz("
digraph advent_goldilocks {
  graph [rankdir = TB, bgcolor = transparent, ranksep = 0.55, nodesep = 0.35]
  node  [shape = box, style = rounded, fontname = Helvetica, fontsize = 14,
         margin = 0.12]
  edge  [fontname = Helvetica, fontsize = 12]

  start [label = 'Start adaptive review\\nN = 350']
  high  [shape = diamond, label = 'Predicted success\\nhigh for both endpoints?']
  success [label = 'Stop enrollment\\ncurrent N is adequate']
  low   [shape = diamond, label = 'Predicted success\\nlow for either endpoint\\nat max N = 750?']
  futile [label = 'Stop enrollment\\nfutility']
  maxn  [shape = diamond, label = 'Reached\\nN = 750?']
  maxstop [label = 'Stop enrollment\\nmaximum reached']
  accrue [label = 'Accrue 100 more\\nmITT subjects']
  follow [label = 'Complete 12-month\\nfollow-up']
  final [label = 'Final Bayesian\\nnoninferiority analyses']

  start -> high
  high -> success [label = 'Yes']
  high -> low [label = 'No']
  low -> futile [label = 'Yes']
  low -> maxn [label = 'No']
  maxn -> maxstop [label = 'Yes']
  maxn -> accrue [label = 'No']
  accrue -> high
  success -> follow
  futile -> follow
  maxstop -> follow
  follow -> final
}
", width = "100%", height = "760px")
```

The corresponding `goldilocks` arguments are:

```{r sample-size-arguments}
N_total <- 750
interim_look <- c(350, 450, 550, 650)

Sn <- c(0.95, 0.90, 0.85, 0.80)
Fn <- c(0.05, 0.10, 0.10, 0.10)
```

The mapping is direct:

-   `N_total = 750` is the maximum modeled mITT analysis-population size.
-   `interim_look = c(350, 450, 550, 650)` supplies the mITT accrual milestones at which the Goldilocks decision rule is evaluated. The maximum sample size is not included in `interim_look`.
-   `Sn` contains the published predicted-success thresholds for sample sizes 350, 450, 550, and 650.
-   `Fn` contains the published futility thresholds for the same looks.

ADVENT required the predictive probability rule to be favorable for both co-primary endpoints before enrollment stopped for predicted success, and unfavorable for either endpoint before stopping for futility. The current `survival_adapt()` interface models one endpoint per run. In this vignette we therefore run the effectiveness and safety endpoints separately, then explain how those separate endpoint-specific runs relate to the reported co-primary design.

The SAP allowed up to 900 enrolled subjects to obtain 750 mITT subjects: up to 750 mITT subjects, 105 nonrandomized roll-in subjects, and 45 randomized subjects who did not enter the mITT population. The package examples below start directly with the analysis population and do not simulate roll-ins or post-randomization exclusions.

## Endpoint scale

ADVENT reported the primary effectiveness endpoint as treatment success at 12 months. The beta-binomial method in `goldilocks` is parameterized on the binary event probability. For the effectiveness endpoint we therefore code the event as failure by 12 months:

$$
p_{\text{failure}} = 1 - p_{\text{treatment success}}.
$$

The design paper reports a target scenario with 65% treatment success in each arm. On the event scale used in the code, that corresponds to a 35% failure probability in each arm.

For the safety endpoint, the event is already an adverse event. The target scenario was an 8% primary safety event rate in each arm.

```{r endpoint-table, echo=FALSE}
endpoint_map <- data.frame(
  Endpoint = c("Effectiveness", "Safety"),
  `Published scale` = c(
    "Treatment success by 12 months",
    "Primary safety event by 12 months"
  ),
  `Code event` = c(
    "Failure to meet treatment success",
    "Primary safety event"
  ),
  `Target event probability` = c(0.35, 0.08),
  `Noninferiority margin` = c(0.15, 0.08),
  `Posterior threshold` = c(0.956, 0.966),
  check.names = FALSE
)

knitr::kable(endpoint_map, digits = 3)
```

The ADVENT primary analyses used Bayesian beta-binomial endpoint models with noninformative $\operatorname{Beta}(0.5, 0.5)$ priors. In `goldilocks`, this is specified with:

```{r beta-prior}
bin_prior <- c(0.5, 0.5)
```

The `prior` argument still appears in the code below because `survival_adapt()` uses a time-to-event model to impute not-yet-observed outcomes at interim looks. That imputation model uses a Gamma prior on piecewise-exponential hazards. The ADVENT SAP specifies a noninformative $\operatorname{Gamma}(0.5, 0.001)$ shape-rate prior for most of those hazards. We use those reported hyperparameters here rather than the package default. Because the SAP expresses the corresponding risk sets and exposure in patient-days, every time supplied to `goldilocks` below is expressed in days:

```{r imputation-prior}
prior <- c(0.5, 0.001)
```

That distinction is important: `bin_prior` is the ADVENT-aligned endpoint prior, whereas `prior` controls the package's predictive imputation model.

The effectiveness model in the SAP needs one further qualification. It partitions follow-up at days 90, 104, 150, and 210. For interim sample-size prediction, the first four interval hazards use $\operatorname{Gamma}(0.5, 0.001)$ priors, but the 210--360-day hazard uses an informative $\operatorname{Gamma}(5, 10000)$ prior because little late follow-up was expected at the interim assessments. The final multiple-imputation analysis returns to $\operatorname{Gamma}(0.5, 0.001)$ for every interval. The current `goldilocks` interface applies one shared Gamma prior to every interval and uses it for both interim prediction and final imputation, so the special 210--360-day prior is not yet available in the package; this limitation is tracked in [GitHub issue #59](https://github.com/graemeleehickey/goldilocks/issues/59). We therefore use $\operatorname{Gamma}(0.5, 0.001)$ throughout this endpoint-specific illustration.

Later follow-up intervals can have no exposure at an early look. The SAP's conjugate model then leaves the corresponding hazard governed by its prior. The package's historical default instead propagates sufficient statistics from a neighboring interval, so the examples explicitly set `empty_interval = "prior"`.

The SAP reports $M = 5000$ completed datasets for its predictive-probability and multiple-imputation calculations. This vignette deliberately uses 30--100 imputations instead: 50 for the worked single trials, 30 for the small operating-characteristic example, and 100 in the unevaluated fuller template. We do not use $M = 5000$ here in order to keep vignette computation time to a minimum. Consequently, the numerical examples are demonstrations of package workflow rather than reproductions of the ADVENT calibration.

## SAP event-time models and time units

The SAP provides the exact default piecewise-exponential generators used in its operating-characteristic simulations. Because their hazards are expressed per patient-day and their Gamma prior rates use patient-day exposure, this vignette uses **days as the single numeric time unit supplied to `goldilocks`**. The SAP also treats 30 days as one simulation month, so 12 months is represented as 360 days.

```{r hazards}
days_per_month <- 30L
follow_up_months <- 12L
end_of_study_day <- follow_up_months * days_per_month

eff_event_cutpoints_day <- c(90, 104, 150, 210)
eff_hazard_per_day <- c(
  0.000111670,
  0.002197976,
  0.003163208,
  0.002839089,
  0.000494053
)

safety_event_cutpoints_day <- 7
safety_hazard_per_day <- c(0.011137363, 1.53540e-5)

hazard_table <- data.frame(
  Endpoint = c(rep("Effectiveness failure", 5), rep("Safety event", 2)),
  `Follow-up interval (days)` = c(
    "0--90", "90--104", "104--150", "150--210", "210--360",
    "0--7", "7--360"
  ),
  `Hazard per patient-day` = c(
    eff_hazard_per_day,
    safety_hazard_per_day
  ),
  check.names = FALSE
)

knitr::kable(hazard_table, digits = 9)
```

The effectiveness hazards imply event-free probabilities of approximately 0.99, 0.96, 0.83, 0.70, and 0.65 at days 90, 104, 150, 210, and 360. The safety hazards imply event-free probabilities of 0.925 at day 7 and 0.920 at day 360. The following calculation verifies the two final binary event probabilities used by the vignette:

```{r probability-check}
prob_check <- data.frame(
  Endpoint = c("Effectiveness failure", "Safety event"),
  `SAP target event probability at day 360` = c(0.35, 0.08),
  `Calculated event probability at day 360` = c(
    ppwe(
      hazard = matrix(eff_hazard_per_day, nrow = 1),
      end_of_study = end_of_study_day,
      cutpoints = eff_event_cutpoints_day
    ),
    ppwe(
      hazard = matrix(safety_hazard_per_day, nrow = 1),
      end_of_study = end_of_study_day,
      cutpoints = safety_event_cutpoints_day
    )
  ),
  check.names = FALSE
)

knitr::kable(prob_check, digits = 6)
```

For scenarios with another day-360 event probability, the SAP multiplies the complete hazard vector by one constant. This preserves the piecewise shape and imposes proportional hazards in data generation. We use the same operation for the noninferiority-boundary examples below:

```{r scaled-hazards}
scale_pwe_to_event_probability <- function(
  hazard_per_day,
  cutpoints_day,
  end_day,
  target_event_probability
) {
  interval_length_day <- diff(c(0, cutpoints_day, end_day))
  cumulative_hazard <- sum(hazard_per_day * interval_length_day)
  scale <- -log1p(-target_event_probability) / cumulative_hazard
  hazard_per_day * scale
}

eff_margin_hazard_per_day <- scale_pwe_to_event_probability(
  eff_hazard_per_day,
  eff_event_cutpoints_day,
  end_of_study_day,
  target_event_probability = 0.50
)

safety_margin_hazard_per_day <- scale_pwe_to_event_probability(
  safety_hazard_per_day,
  safety_event_cutpoints_day,
  end_of_study_day,
  target_event_probability = 0.16
)
```

Two time clocks remain conceptually distinct even though both are represented in days. Event cut-points are subject-relative follow-up times; enrollment-rate change points below are trial-calendar times measured from first patient in. In addition, the SAP's event clock starts on the index-procedure date, whereas `goldilocks` starts it at enrollment/randomization. This vignette treats those dates as coincident and does not model a randomization-to-procedure delay.

## Effectiveness endpoint

For the effectiveness endpoint, success in the code means:

$$
\Pr(p_{\text{PFA failure}} - p_{\text{thermal failure}} < 0.15 \mid
\text{data}) > 0.956.
$$

This matches the reported 15 percentage point absolute noninferiority margin for treatment success, after translating success into failure. A lower failure probability is better, so we use `alternative = "less"`.

ADVENT used 1:1 randomization stratified by site. The design paper reports randomly varying permuted block sizes, but the SAP does not provide the block sizes or the randomization-generation algorithm. The package can generate blocked 1:1 randomization, but it does not model site-level stratification or randomly varying block sizes. Here `block = 2` and `rand_ratio = c(1, 1)` retain simple 1:1 balance; they do not reproduce the trial's full randomization scheme.

The SAP explicitly designates 5% of subjects as lost to follow-up for safety and 7.5% for effectiveness: the same 5% plus another 2.5% who may be unassessable for effectiveness while remaining assessable for safety. SAP dropout times are uniform over days 0--360, and an event remains observed when it precedes dropout. The package's `prop_loss` mechanism is similar but not identical: every selected subject is censored at a uniformly generated time before their potential event or administrative censoring time. Separate endpoint runs also cannot preserve the SAP's shared subject-level 5% plus nested 2.5% relationship. We therefore use 0.075 and 0.05 to map the SAP's **designated fractions**, not to claim exact reproduction of its censoring process. The subsequently observed proportion not completing follow-up was approximately 0.04, which is close to 0.05, but that post-trial quantity is not the design input and need not equal endpoint-specific missingness.

The SAP reports the expected accrual schedule in subjects per month: 2, 5, 10, 15, 20, 25, and 30 during months 1--7, then 33 per month from month 8 onward. Accrual is into the primary analysis population. The source rates remain visibly labeled per month below, but they are divided by 30 exactly once before being supplied to `goldilocks`. Likewise, the calendar-month change points are converted once to days since first patient in. The SAP does not describe its individual arrival-time generator; `goldilocks` uses these rates in a continuous-time Poisson enrollment process with first patient in fixed at calendar day 0.

```{r accrual-and-missingness}
enrollment_rate_per_month <- c(2, 5, 10, 15, 20, 25, 30, 33)
enrollment_rate_change_month <- 1:7

enrollment_rate_per_day <- enrollment_rate_per_month / days_per_month
enrollment_rate_change_day <-
  enrollment_rate_change_month * days_per_month

effectiveness_prop_loss <- 0.075
safety_prop_loss <- 0.05

enrollment_schedule <- data.frame(
  `Trial-calendar interval` = c(
    paste("Month", 1:7),
    "Month 8 onward"
  ),
  `SAP rate (subjects/month)` = enrollment_rate_per_month,
  `Rate supplied to goldilocks (subjects/day)` = enrollment_rate_per_day,
  check.names = FALSE
)

knitr::kable(enrollment_schedule, digits = 4)
```

Thus, for example, trial-calendar day 210 is the beginning of month 8 and the 33-subject/month rate. It is unrelated to the effectiveness endpoint's subject-relative cut-point at follow-up day 210. The 360-day follow-up horizon is also per subject, not the total duration of the trial. Accrual speed remains worth sensitivity checking: faster accrual reduces the amount of observed endpoint information available at interim looks.

```{r effectiveness-fit}
set.seed(4601)

advent_effectiveness <- survival_adapt(
  hazard_treatment = eff_hazard_per_day,
  hazard_control = eff_hazard_per_day,
  cutpoints = eff_event_cutpoints_day,
  N_total = N_total,
  lambda = enrollment_rate_per_day,
  lambda_time = enrollment_rate_change_day,
  interim_look = interim_look,
  end_of_study = end_of_study_day,
  prior = prior,
  bin_prior = bin_prior,
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = effectiveness_prop_loss,
  alternative = "less",
  h0 = 0.15,
  Fn = Fn,
  Sn = Sn,
  prob_ha = 0.956,
  N_impute = 50,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE
)

advent_effectiveness
```

The most important columns are:

-   `N_enrolled`: the selected sample size for this simulated trial.
-   `stop_expected_success`: whether enrollment stopped because the current sample size appeared adequate.
-   `stop_futility`: whether enrollment stopped because success looked unlikely even at the maximum sample size.
-   `est_final`: the posterior mean of $p_{\text{PFA failure}} - p_{\text{thermal failure}}$.
-   `post_prob_ha`: the final posterior probability that the binary endpoint effect is below the noninferiority margin.

## Safety endpoint

The safety endpoint uses the same Goldilocks sample-size rule, but changes the event probability, margin, and posterior probability threshold. Success in the code means:

$$
\Pr(p_{\text{PFA safety event}} - p_{\text{thermal safety event}} < 0.08
\mid \text{data}) > 0.966.
$$

```{r safety-fit}
set.seed(4602)

advent_safety <- survival_adapt(
  hazard_treatment = safety_hazard_per_day,
  hazard_control = safety_hazard_per_day,
  cutpoints = safety_event_cutpoints_day,
  N_total = N_total,
  lambda = enrollment_rate_per_day,
  lambda_time = enrollment_rate_change_day,
  interim_look = interim_look,
  end_of_study = end_of_study_day,
  prior = prior,
  bin_prior = bin_prior,
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = safety_prop_loss,
  alternative = "less",
  h0 = 0.08,
  Fn = Fn,
  Sn = Sn,
  prob_ha = 0.966,
  N_impute = 50,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE
)

advent_safety
```

In the published ADVENT design, a predicted-success stopping recommendation required high predictive probability for both endpoints. A futility recommendation could be triggered by low predictive probability for either endpoint. The two separate simulations above do not impose that joint rule. Instead, they show how each endpoint-specific Bayesian rule maps to `survival_adapt()`.

## A compact design object

For simulations, it is helpful to collect the settings that genuinely are common and then add endpoint-specific cut-points and designated missingness fractions. Keeping those endpoint-specific arguments out of the common object prevents an effectiveness setting from being reused accidentally for safety, or vice versa.

```{r common-design}
advent_common <- list(
  N_total = N_total,
  lambda = enrollment_rate_per_day,
  lambda_time = enrollment_rate_change_day,
  interim_look = interim_look,
  end_of_study = end_of_study_day,
  prior = prior,
  bin_prior = bin_prior,
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  alternative = "less",
  Fn = Fn,
  Sn = Sn,
  N_impute = 30,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE,
  ncores = 1
)

advent_effectiveness_args <- modifyList(advent_common, list(
  cutpoints = eff_event_cutpoints_day,
  prop_loss = effectiveness_prop_loss
))

advent_safety_args <- modifyList(advent_common, list(
  cutpoints = safety_event_cutpoints_day,
  prop_loss = safety_prop_loss
))

advent_effectiveness_args
```

Then the endpoint-specific simulation calls become short:

```{r oc-small}
set.seed(4610)

eff_target <- do.call(sim_trials, c(
  advent_effectiveness_args,
  list(
    N_trials = 20,
    hazard_treatment = eff_hazard_per_day,
    hazard_control = eff_hazard_per_day,
    h0 = 0.15,
    prob_ha = 0.956,
    return_trace = TRUE,
    seed = 4610
  )
))

eff_null_boundary <- do.call(sim_trials, c(
  advent_effectiveness_args,
  list(
    N_trials = 20,
    hazard_treatment = eff_margin_hazard_per_day,
    hazard_control = eff_hazard_per_day,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4611
  )
))

oc_small <- summarise_sims(list(
  "target: equal 35% failure" = eff_target$sims,
  "margin: PFA failure 50%" = eff_null_boundary$sims
))

knitr::kable(oc_small, digits = 3)
```

This table is intentionally small so the vignette can build quickly. It checks that the code runs and shows the workflow for comparing a target scenario with a boundary scenario. It should not be interpreted as a stable estimate of power or type I error. For design work, increase `N_trials`, increase `N_impute`, and run sensitivity analyses over accrual speed, loss to follow-up, event rates, and the imputation hazard model.

## Visualizing the simulated design

The plotting helpers answer complementary design questions. First,
`plot_sim_ocs()` compares operating characteristics across the target and
noninferiority-margin scenarios. The true PFA event probability is supplied as
the effect scale.

```{r advent-oc-plot, fig.width=9, fig.height=4.5}
oc_small$true_pfa_event_probability <- c(0.35, 0.50)
plot_sim_ocs(
  oc_small,
  effect = "true_pfa_event_probability",
  xlab = "True 12-month PFA event probability"
)
```

Within the target scenario, `plot_sim_stopping()` shows how often each sample
size is selected and why enrollment stops.

```{r advent-stopping-plot, fig.width=7, fig.height=5}
plot_sim_stopping(eff_target)
```

Because `eff_target` was simulated with `return_trace = TRUE`, its interim
decision geometry can also be inspected. Each displayed panel corresponds to
an ADVENT enrollment milestone reached in at least one simulated trial; larger
bubbles represent predictive-probability pairs reached by more simulated
trials.

```{r advent-decision-plot, fig.width=9, fig.height=8}
plot_sim_decisions(eff_target)
```

## A fuller simulation template

The following code is closer to what one would run outside the vignette build. It is not evaluated here.

```{r full-calibration-template, eval=FALSE}
advent_effectiveness_full_args <- modifyList(advent_effectiveness_args, list(
  N_impute = 100,
  ncores = 8
))

advent_safety_full_args <- modifyList(advent_safety_args, list(
  N_impute = 100,
  ncores = 8
))

eff_target_full <- do.call(sim_trials, c(
  advent_effectiveness_full_args,
  list(
    N_trials = 1000,
    hazard_treatment = eff_hazard_per_day,
    hazard_control = eff_hazard_per_day,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4620
  )
))

eff_margin_full <- do.call(sim_trials, c(
  advent_effectiveness_full_args,
  list(
    N_trials = 1000,
    hazard_treatment = eff_margin_hazard_per_day,
    hazard_control = eff_hazard_per_day,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4621
  )
))

safety_target_full <- do.call(sim_trials, c(
  advent_safety_full_args,
  list(
    N_trials = 1000,
    hazard_treatment = safety_hazard_per_day,
    hazard_control = safety_hazard_per_day,
    h0 = 0.08,
    prob_ha = 0.966,
    seed = 4622
  )
))

safety_margin_full <- do.call(sim_trials, c(
  advent_safety_full_args,
  list(
    N_trials = 1000,
    hazard_treatment = safety_margin_hazard_per_day,
    hazard_control = safety_hazard_per_day,
    h0 = 0.08,
    prob_ha = 0.966,
    seed = 4623
  )
))

summarise_sims(list(
  "effectiveness target" = eff_target_full$sims,
  "effectiveness margin" = eff_margin_full$sims,
  "safety target" = safety_target_full$sims,
  "safety margin" = safety_margin_full$sims
))
```

The "margin" scenarios above set the treatment event probability equal to the control event probability plus the noninferiority margin. They are useful for checking false-positive behavior near the design boundary. They are not the only null scenarios worth studying.

## SAP sensitivity assumptions and reference benchmarks

The SAP assessed whether operating characteristics were robust to several alternative assumptions:

-   Control effectiveness success was varied over $P_C \in \{0.60, 0.65, 0.70\}$ and the control safety-event rate over $Q_C \in \{0.06, 0.08, 0.10\}$.
-   Slower accrual used 2, 5, 10, 10, 15, 15, and 20 subjects/month during months 1--7, then 25/month. Faster accrual used 5, 10, 20, 25, 30, and 40 subjects/month during months 1--6, then 50/month.
-   Effectiveness generator Eff2 used cut-points at days 90, 100, 200, and 300 with daily hazards `c(0.000055695, 0.010034797, 0.001690763, 0.000680535, 0.000573122)`.
-   Effectiveness generator Eff3 used cut-points at days 90 and 120 with daily hazards `c(0.000569925, 0.007879626, 0.000596254)`.
-   Safety generator Saf2 used cut-points at days 7 and 180 with daily hazards `c(0.007327613, 0.000122991, 0.0000600610)`.

Eff2, Eff3, and Saf2 deliberately use data-generating partitions that differ from the principal analysis partitions. The current package accepts one `cutpoints` vector for both event generation and predictive imputation, so it cannot reproduce those model-misspecification sensitivities directly. The SAP also planned an imputation-model sensitivity with six pieces: day 90 followed by five post-90-day intervals containing approximately equal event counts, potentially with different cut-points by arm. A shared fixed package cut-point vector cannot reproduce that analysis either.

The SAP supplies quantitative reference results based on 10,000 simulated trials per scenario and $M=5000$ predictive draws or imputations:

```{r sap-reference-benchmarks, echo=FALSE}
sap_reference_benchmarks <- data.frame(
  Scenario = c(
    "Target: both endpoints at expected rates",
    "Effectiveness noninferiority boundary",
    "Safety noninferiority boundary"
  ),
  `Reported probability` = c(0.9635, 0.0487, 0.0528),
  `Probability meaning` = c(
    "Joint power",
    "Effectiveness type I error",
    "Safety type I error"
  ),
  `Mean mITT sample size` = c(504, 551, 457),
  check.names = FALSE
)

knitr::kable(sap_reference_benchmarks, digits = 4)
```

The target row assumes $P_T=P_C=0.65$ and $Q_T=Q_C=0.08$, with success requiring both co-primary endpoints. At the effectiveness boundary, $P_T=0.50$ and the safety endpoint was assumed always to recommend stopping for promise and to pass finally. At the safety boundary, $Q_T=0.16$ and effectiveness was treated analogously. These are external validation targets. The small, endpoint-separated simulations in this vignette do not claim to reproduce them.

## What matches ADVENT, and what is simplified?

The code above is intentionally close to the reported design in the following ways:

-   It uses the published Goldilocks sample-size options: 350, 450, 550, 650, and 750.
-   It uses the published predicted-success thresholds: 0.95, 0.90, 0.85, and 0.80.
-   It uses the published futility thresholds: 0.05, 0.10, 0.10, and 0.10.
-   It uses the published beta-binomial endpoint prior, $\operatorname{Beta}(0.5, 0.5)$, via `bin_prior = c(0.5, 0.5)`.
-   It uses the published noninferiority margins: 15 percentage points for effectiveness and 8 percentage points for safety.
-   It uses the published posterior probability thresholds: 0.956 for effectiveness and 0.966 for safety.
-   It uses the SAP's default Eff1 and Saf1 piecewise event-time generators, including their day-scale cut-points and daily hazards.
-   It uses a 360-day representation of the 12-month endpoint window, keeping the time unit consistent with the SAP's Gamma prior rates and interval definitions.
-   It uses the SAP's exact expected accrual ramp after converting the reported subjects/month rates to subjects/day for the day-scale package simulation.
-   It maps the SAP's designated missingness fractions of 7.5% for effectiveness and 5% for safety.
-   It uses the SAP's reported $\operatorname{Gamma}(0.5, 0.001)$ imputation-prior hyperparameters where the current package interface permits a shared prior, and leaves empty intervals prior-driven with `empty_interval = "prior"`.

The code simplifies the reported trial in these ways:

-   It models effectiveness and safety as separate endpoint-specific `survival_adapt()` runs. ADVENT used co-primary endpoints with a joint stopping recommendation.
-   It models 350--750 analysis-population subjects directly and does not simulate roll-ins, randomized subjects excluded from mITT, or the difference between the mITT effectiveness and Safety Populations.
-   It does not model site stratification or the randomly varying block randomization scheme; `block = 2` is an illustrative package setting.
-   It cannot reproduce the SAP dropout mechanism or the shared subject-level missingness relationship across endpoints, even though the designated marginal fractions are supplied.
-   It cannot apply the effectiveness endpoint's special $\operatorname{Gamma}(5, 10000)$ prior to the 210--360-day interval during interim prediction and then switch that interval to $\operatorname{Gamma}(0.5, 0.001)$ for final imputation.
-   It cannot use different data-generation and analysis cut-points for the Eff2, Eff3, and Saf2 sensitivity generators, or arm-specific data-driven cut-points for the six-piece imputation sensitivity.
-   It treats enrollment/randomization and the index procedure as coincident and does not reproduce endpoint-specific monitoring and censoring dates.
-   It uses 30--100 imputations for tractable vignette computation, whereas ADVENT used $M = 5000$.

The SAP now provides exact inputs and reference operating characteristics, so it supports a stronger, documented component-level validation exercise than the design paper alone. The comparison supports validation of the endpoint-level mechanics that map directly to `goldilocks`; it does not validate the package as a complete reproduction of ADVENT. A full reproduction would additionally require the joint co-primary stopping and final-success rules, the SAP censoring and shared-missingness process, interval- and stage-specific priors, and sufficiently large simulations compared directly with the published benchmarks.

## References

Broglio KR, Connor JT, Berry SM. Not too big, not too small: a Goldilocks approach to sample size selection. *Journal of Biopharmaceutical Statistics*. 2014;24(3):685-705. doi:10.1080/10543406.2014.888569.

Reddy VY, Gerstenfeld EP, Natale A, et al. A randomized controlled trial of pulsed field ablation versus standard-of-care ablation for paroxysmal atrial fibrillation: The ADVENT trial rationale and design. *Heart Rhythm O2*. 2023;4(5):317-328. doi:10.1016/j.hroo.2023.03.001.

Reddy VY, Gerstenfeld EP, Natale A, et al. Pulsed field or conventional thermal ablation for paroxysmal atrial fibrillation. *New England Journal of Medicine*. 2023;389(18):1660-1671. doi:10.1056/NEJMoa2307291.

ClinicalTrials.gov. A prospective randomized pivotal trial of the FARAPULSE pulsed field ablation system compared with standard of care ablation in patients with paroxysmal atrial fibrillation. NCT04612244.

FARAPULSE, Inc. *ADVENT Statistical Analysis Plan*. CS0935 Rev C. 12 October 2021. Available from the [ClinicalTrials.gov document archive](https://cdn.clinicaltrials.gov/large-docs/44/NCT04612244/SAP_001.pdf).
