---
title: "ADVENT: a published Goldilocks design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ADVENT: a published Goldilocks design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "advent-cache/",
  fig.width = 7,
  fig.height = 4
)
set.seed(46012244)
```

```{r setup, message=FALSE}
library(goldilocks)
```

The ADVENT trial is a useful published example of a Goldilocks adaptive sample
size design. ADVENT compared pulsed field ablation (PFA) with conventional
thermal ablation for patients with drug-resistant paroxysmal atrial
fibrillation. It was registered as
[NCT04612244](https://clinicaltrials.gov/study/NCT04612244), its design was
published in *Heart Rhythm O2*, and its primary results were published in
*The New England Journal of Medicine*.

The design paper states that the randomized sample size was determined
adaptively using a Goldilocks design, citing Broglio, Connor, and Berry (2014).
The possible randomized sample sizes were 350, 450, 550, 650, and 750. At each
enrollment milestone, the trial calculated the predictive probability that the
trial would eventually show noninferiority for both primary endpoints. The
trial could then stop enrollment for predicted success, stop for futility, or
continue to the next milestone.

This vignette uses ADVENT as a worked example for `goldilocks`. It focuses on
how the published design maps to package arguments. The goal is not to recreate
the sponsor's full statistical analysis plan exactly. Some details, including
site-stratified randomization, the exact composite-event data-generating
process, and the joint co-primary endpoint stopping rule, are simplified here.

## Trial overview

ADVENT was a multicenter, prospective, single-blind, randomized controlled
noninferiority trial. Randomized subjects were assigned 1:1 to PFA or
standard-of-care thermal ablation, where the thermal arm used either
radiofrequency ablation or cryoballoon ablation depending on site. The first 1
to 3 subjects at each site were nonrandomized roll-in subjects and are not part
of the randomized comparison modeled below.

```{r trial-summary}
trial_summary <- data.frame(
  Feature = c(
    "Population",
    "Treatment arm",
    "Control arm",
    "Randomization",
    "Follow-up",
    "Adaptive sample sizes",
    "Primary effectiveness endpoint",
    "Primary safety endpoint"
  ),
  Reported_ADVENT_design = c(
    "Drug-resistant paroxysmal atrial fibrillation",
    "Pulsed field ablation",
    "Thermal ablation by radiofrequency or cryoballoon ablation",
    "1:1 after nonrandomized roll-in subjects",
    "12 months",
    "350, 450, 550, 650, or 750 randomized subjects",
    paste(
      "Treatment success: acute procedural success and freedom from",
      "specified chronic failures through 12 months"
    ),
    paste(
      "Composite device- or procedure-related serious adverse events,",
      "including selected acute and chronic events"
    )
  )
)

knitr::kable(trial_summary)
```

The final published randomized cohort included 305 subjects assigned to PFA and
302 assigned to thermal ablation. In the primary results paper, PFA was
noninferior to thermal ablation for both primary effectiveness and primary
safety.

The published design also included a trial-flow figure. The following diagram
recreates the parts of that flow that matter for the package mapping.

```{r trial-flowchart, message=FALSE}
DiagrammeR::grViz("
digraph advent_trial_flow {
  graph [rankdir = TB, bgcolor = transparent]
  node  [shape = box, style = rounded, fontname = Helvetica, fontsize = 10]
  edge  [fontname = Helvetica, fontsize = 9]

  screen [label = 'Eligible patients with\\ndrug-resistant paroxysmal AF']
  rollin [label = 'Optional site roll-in\\n1 to 3 nonrandomized PFA subjects']
  random [label = 'Randomized subjects\\n1:1 allocation']
  pfa    [label = 'PFA arm\\nFARAPULSE system']
  thermal [label = 'Thermal ablation arm\\nradiofrequency or cryoballoon']
  blank  [label = '90-day blanking period\\nfor arrhythmia recurrence']
  follow [label = 'Follow-up through\\n12 months']
  endpoints [label = 'Co-primary endpoints\\neffectiveness and safety']

  screen -> rollin
  rollin -> random
  random -> pfa
  random -> thermal
  pfa -> blank
  thermal -> blank
  blank -> follow
  follow -> endpoints
}
")
```

In the code below, the nonrandomized roll-in subjects are ignored and the
randomized comparison starts at the 1:1 allocation step.

## The Goldilocks flow

The ADVENT design paper includes a flowchart for the adaptive sample size
algorithm. The chart below recreates the logic in package terms rather than
copying the published image.

```{r adaptive-flowchart, message=FALSE}
DiagrammeR::grViz("
digraph advent_goldilocks {
  graph [rankdir = LR, bgcolor = transparent]
  node  [shape = box, style = rounded, fontname = Helvetica, fontsize = 10]
  edge  [fontname = Helvetica, fontsize = 9]

  start [label = 'Enroll N = 350\\nrandomized subjects']
  high  [shape = diamond, label = 'High predictive probability\\nof noninferiority for\\nboth endpoints at current N?']
  success [label = 'Stop enrollment\\nSample size is adequate']
  low   [shape = diamond, label = 'Low predictive probability\\nof noninferiority for\\neither endpoint at max N = 750?']
  futile [label = 'Stop enrollment\\nTrial appears futile']
  maxn  [shape = diamond, label = 'N = 750?']
  maxstop [label = 'Stop enrollment\\nMaximum reached']
  accrue [label = 'Accrue 100 more\\nrandomized subjects']
  follow [label = 'Follow enrolled subjects\\nto 12 months']
  final [label = 'Final Bayesian\\nnoninferiority analyses']

  start -> high
  high -> success [label = 'Yes']
  high -> low [label = 'No']
  low -> futile [label = 'Yes']
  low -> maxn [label = 'No']
  maxn -> maxstop [label = 'Yes']
  maxn -> accrue [label = 'No']
  accrue -> high
  success -> follow
  futile -> follow
  maxstop -> follow
  follow -> final
}
")
```

The corresponding `goldilocks` arguments are:

```{r sample-size-arguments}
N_total <- 750
interim_look <- c(350, 450, 550, 650)

Sn <- c(0.95, 0.90, 0.85, 0.80)
Fn <- c(0.05, 0.10, 0.10, 0.10)
```

The mapping is direct:

- `N_total = 750` is the maximum randomized sample size.
- `interim_look = c(350, 450, 550, 650)` supplies the enrollment milestones at
  which the Goldilocks decision rule is evaluated. The maximum sample size is
  not included in `interim_look`.
- `Sn` contains the published predicted-success thresholds for sample sizes
  350, 450, 550, and 650.
- `Fn` contains the published futility thresholds for the same looks.

ADVENT required the predictive probability rule to be favorable for both
co-primary endpoints before enrollment stopped for predicted success, and
unfavorable for either endpoint before stopping for futility. The current
`survival_adapt()` interface models one endpoint per run. In this vignette we
therefore run the effectiveness and safety endpoints separately, then explain
how those separate endpoint-specific runs relate to the reported co-primary
design.

## Endpoint scale

ADVENT reported the primary effectiveness endpoint as treatment success at 12
months. The beta-binomial method in `goldilocks` is parameterized on the binary
event probability. For the effectiveness endpoint we therefore code the event as
failure by 12 months:

$$
p_{\text{failure}} = 1 - p_{\text{treatment success}}.
$$

The design paper reports a target scenario with 65% treatment success in each
arm. On the event scale used in the code, that corresponds to a 35% failure
probability in each arm.

For the safety endpoint, the event is already an adverse event. The target
scenario was an 8% primary safety event rate in each arm.

```{r endpoint-table}
endpoint_map <- data.frame(
  Endpoint = c("Effectiveness", "Safety"),
  Published_scale = c(
    "Treatment success by 12 months",
    "Primary safety event by 12 months"
  ),
  Code_event = c(
    "Failure to meet treatment success",
    "Primary safety event"
  ),
  Target_event_probability = c(0.35, 0.08),
  Noninferiority_margin = c(0.15, 0.08),
  Posterior_threshold = c(0.956, 0.966)
)

knitr::kable(endpoint_map, digits = 3)
```

The ADVENT primary analyses used Bayesian beta-binomial endpoint models with
noninformative `Beta(0.5, 0.5)` priors. In `goldilocks`, this is specified with:

```{r beta-prior}
bin_prior <- c(0.5, 0.5)
```

The `prior` argument still appears in the code below because `survival_adapt()`
uses a time-to-event model to impute not-yet-observed outcomes at interim looks.
That imputation model uses a Gamma prior on piecewise-exponential hazards. The
ADVENT design paper reports the beta-binomial prior for the completed binary
endpoint analysis, but the main paper does not provide the exact Gamma
hyperparameters used for the time-to-event predictive imputation model. We
therefore use the package default:

```{r imputation-prior}
prior <- c(0.1, 0.1)
```

That distinction is important: `bin_prior` is the ADVENT-aligned endpoint prior;
`prior` is the package's predictive imputation prior.

## Simulating endpoint probabilities

The `goldilocks` simulator generates event times and then reduces each imputed
or completed dataset to binary endpoint status when `method = "bayes-bin"`.
For a simple vignette, we use constant hazards that reproduce the target event
probabilities at 12 months.

```{r hazards}
end_of_study <- 12

haz_eff_fail <- prop_to_haz(0.35, endtime = end_of_study)
haz_safety <- prop_to_haz(0.08, endtime = end_of_study)

prob_check <- data.frame(
  Endpoint = c("Effectiveness failure", "Safety event"),
  Hazard = c(haz_eff_fail, haz_safety),
  Event_probability_at_12_months = c(
    ppwe(
      hazard = matrix(haz_eff_fail, nrow = 1),
      end_of_study = end_of_study,
      cutpoints = 0
    ),
    ppwe(
      hazard = matrix(haz_safety, nrow = 1),
      end_of_study = end_of_study,
      cutpoints = 0
    )
  )
)

knitr::kable(prob_check, digits = 4)
```

This constant-hazard simplification is a modeling choice for the vignette. The
ADVENT composite endpoints combine acute procedural outcomes and later chronic
outcomes. A richer analysis could use piecewise hazards to place more event
mass early, for example around the index procedure or shortly after the
blanking period. The package supports that through `cutpoints` and vector-valued
hazards, but the published design summary is sufficient for the simpler
constant-hazard example here.

## Effectiveness endpoint

For the effectiveness endpoint, success in the code means:

$$
\Pr(p_{\text{PFA failure}} - p_{\text{thermal failure}} < 0.15 \mid
\text{data}) > 0.956.
$$

This matches the reported 15 percentage point absolute noninferiority margin
for treatment success, after translating success into failure. A lower failure
probability is better, so we use `alternative = "less"`.

ADVENT used 1:1 randomization. The design paper also reports site
stratification and randomly varying permuted block sizes. The package can
generate blocked 1:1 randomization, but it does not model site-level
stratification. Here `block = 2` and `rand_ratio = c(1, 1)` represent the 1:1
allocation.

The paper reports follow-up through 12 months. The ClinicalTrials.gov record
reports 305 randomized subjects starting in the PFA arm and 302 in the thermal
arm, with 12 not completing in each arm. We use `prop_loss = 0.04` as a rough
vignette-level approximation to incomplete follow-up.

The reported enrollment period implies rapid accrual, roughly 40 randomized
subjects per month. We use `lambda = 40` to make the simulated calendar-time
information broadly comparable. Accrual speed is worth sensitivity checking:
fast accrual can reduce the amount of observed endpoint information available
at interim looks.

```{r effectiveness-fit}
set.seed(4601)

advent_effectiveness <- survival_adapt(
  hazard_treatment = haz_eff_fail,
  hazard_control = haz_eff_fail,
  cutpoints = 0,
  N_total = N_total,
  lambda = 40,
  lambda_time = 0,
  interim_look = interim_look,
  end_of_study = end_of_study,
  prior = prior,
  bin_prior = bin_prior,
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0.04,
  alternative = "less",
  h0 = 0.15,
  Fn = Fn,
  Sn = Sn,
  prob_ha = 0.956,
  N_impute = 50,
  method = "bayes-bin",
  imputed_final = TRUE
)

advent_effectiveness
```

The most important columns are:

- `N_enrolled`: the selected sample size for this simulated trial.
- `stop_expected_success`: whether enrollment stopped because the current
  sample size appeared adequate.
- `stop_futility`: whether enrollment stopped because success looked unlikely
  even at the maximum sample size.
- `est_final`: the posterior mean of
  \(p_{\text{PFA failure}} - p_{\text{thermal failure}}\).
- `post_prob_ha`: the final posterior probability that the binary endpoint
  effect is below the noninferiority margin.

## Safety endpoint

The safety endpoint uses the same Goldilocks sample-size rule, but changes the
event probability, margin, and posterior probability threshold. Success in the
code means:

$$
\Pr(p_{\text{PFA safety event}} - p_{\text{thermal safety event}} < 0.08
\mid \text{data}) > 0.966.
$$

```{r safety-fit}
set.seed(4602)

advent_safety <- survival_adapt(
  hazard_treatment = haz_safety,
  hazard_control = haz_safety,
  cutpoints = 0,
  N_total = N_total,
  lambda = 40,
  lambda_time = 0,
  interim_look = interim_look,
  end_of_study = end_of_study,
  prior = prior,
  bin_prior = bin_prior,
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0.04,
  alternative = "less",
  h0 = 0.08,
  Fn = Fn,
  Sn = Sn,
  prob_ha = 0.966,
  N_impute = 50,
  method = "bayes-bin",
  imputed_final = TRUE
)

advent_safety
```

In the published ADVENT design, a predicted-success stopping recommendation
required high predictive probability for both endpoints. A futility
recommendation could be triggered by low predictive probability for either
endpoint. The two separate simulations above do not impose that joint rule.
Instead, they show how each endpoint-specific Bayesian rule maps to
`survival_adapt()`.

## A compact design object

For simulations, it is helpful to collect the common design settings once and
then vary only the endpoint-specific values.

```{r common-design}
advent_common <- list(
  cutpoints = 0,
  N_total = N_total,
  lambda = 40,
  lambda_time = 0,
  interim_look = interim_look,
  end_of_study = end_of_study,
  prior = prior,
  bin_prior = bin_prior,
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0.04,
  alternative = "less",
  Fn = Fn,
  Sn = Sn,
  N_impute = 30,
  method = "bayes-bin",
  imputed_final = TRUE,
  ncores = 1
)

advent_common
```

Then the endpoint-specific simulation calls become short:

```{r oc-small}
set.seed(4610)

eff_target <- do.call(sim_trials, c(
  advent_common,
  list(
    N_trials = 20,
    hazard_treatment = haz_eff_fail,
    hazard_control = haz_eff_fail,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4610
  )
))

eff_null_boundary <- do.call(sim_trials, c(
  advent_common,
  list(
    N_trials = 20,
    hazard_treatment = prop_to_haz(0.50, endtime = end_of_study),
    hazard_control = haz_eff_fail,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4611
  )
))

oc_small <- summarise_sims(list(
  "target: equal 35% failure" = eff_target$sims,
  "margin: PFA failure 50%" = eff_null_boundary$sims
))

knitr::kable(oc_small, digits = 3)
```

This table is intentionally small so the vignette can build quickly. It checks
that the code runs and shows the workflow for comparing a target scenario with
a boundary scenario. It should not be interpreted as a stable estimate of power
or type I error. For design work, increase `N_trials`, increase `N_impute`, and
run sensitivity analyses over accrual speed, loss to follow-up, event rates,
and the imputation hazard model.

## Full calibration template

The following code is closer to what one would run outside the vignette build.
It is not evaluated here.

```{r full-calibration-template, eval=FALSE}
advent_common_full <- modifyList(advent_common, list(
  N_impute = 100,
  ncores = 8
))

eff_target_full <- do.call(sim_trials, c(
  advent_common_full,
  list(
    N_trials = 1000,
    hazard_treatment = haz_eff_fail,
    hazard_control = haz_eff_fail,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4620
  )
))

eff_margin_full <- do.call(sim_trials, c(
  advent_common_full,
  list(
    N_trials = 1000,
    hazard_treatment = prop_to_haz(0.50, endtime = end_of_study),
    hazard_control = haz_eff_fail,
    h0 = 0.15,
    prob_ha = 0.956,
    seed = 4621
  )
))

safety_target_full <- do.call(sim_trials, c(
  advent_common_full,
  list(
    N_trials = 1000,
    hazard_treatment = haz_safety,
    hazard_control = haz_safety,
    h0 = 0.08,
    prob_ha = 0.966,
    seed = 4622
  )
))

safety_margin_full <- do.call(sim_trials, c(
  advent_common_full,
  list(
    N_trials = 1000,
    hazard_treatment = prop_to_haz(0.16, endtime = end_of_study),
    hazard_control = haz_safety,
    h0 = 0.08,
    prob_ha = 0.966,
    seed = 4623
  )
))

summarise_sims(list(
  "effectiveness target" = eff_target_full$sims,
  "effectiveness margin" = eff_margin_full$sims,
  "safety target" = safety_target_full$sims,
  "safety margin" = safety_margin_full$sims
))
```

The "margin" scenarios above set the treatment event probability equal to the
control event probability plus the noninferiority margin. They are useful for
checking false-positive behavior near the design boundary. They are not the only
null scenarios worth studying.

## What matches ADVENT, and what is simplified?

The code above is intentionally close to the reported design in the following
ways:

- It uses the published Goldilocks sample-size options: 350, 450, 550, 650,
  and 750.
- It uses the published predicted-success thresholds: 0.95, 0.90, 0.85, and
  0.80.
- It uses the published futility thresholds: 0.05, 0.10, 0.10, and 0.10.
- It uses the published beta-binomial endpoint prior, `Beta(0.5, 0.5)`, via
  `bin_prior = c(0.5, 0.5)`.
- It uses the published noninferiority margins: 15 percentage points for
  effectiveness and 8 percentage points for safety.
- It uses the published posterior probability thresholds: 0.956 for
  effectiveness and 0.966 for safety.
- It uses a 12-month endpoint window.

The code simplifies the reported trial in these ways:

- It models effectiveness and safety as separate endpoint-specific
  `survival_adapt()` runs. ADVENT used co-primary endpoints with a joint
  stopping recommendation.
- It does not model site stratification or the exact randomly varying block
  randomization scheme.
- It ignores nonrandomized roll-in subjects.
- It uses constant hazards to reproduce the target 12-month event
  probabilities. The real composite endpoints include acute and chronic
  components.
- It uses the package default Gamma prior for the time-to-event imputation
  model because the main design paper reports the beta-binomial endpoint prior
  but not all imputation-model hyperparameters.
- It approximates incomplete follow-up with `prop_loss = 0.04`.

These simplifications are typical when translating a published design into a
teaching vignette. The important point is that the central ADVENT design
features now have direct `goldilocks` arguments.

## References

Broglio KR, Connor JT, Berry SM. Not too big, not too small: a Goldilocks
approach to sample size selection. *Journal of Biopharmaceutical Statistics*.
2014;24(3):685-705. doi:10.1080/10543406.2014.888569.

Reddy VY, Gerstenfeld EP, Natale A, et al. A randomized controlled trial of
pulsed field ablation versus standard-of-care ablation for paroxysmal atrial
fibrillation: The ADVENT trial rationale and design. *Heart Rhythm O2*.
2023;4(5):317-328. doi:10.1016/j.hroo.2023.03.001.

Reddy VY, Gerstenfeld EP, Natale A, et al. Pulsed field or conventional thermal
ablation for paroxysmal atrial fibrillation. *New England Journal of Medicine*.
2023;389(18):1660-1671. doi:10.1056/NEJMoa2307291.

ClinicalTrials.gov. A prospective randomized pivotal trial of the FARAPULSE
pulsed field ablation system compared with standard of care ablation in
patients with paroxysmal atrial fibrillation. NCT04612244.
