---
title: "Package architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `goldilocks` package has three ways into the adaptive calculation:

1.  `sim_trials()` simulates many independent trials and delegates each one to `survival_adapt()`;
2.  `survival_adapt()` generates and evaluates one complete simulated trial; and
3.  `evaluate_interim()` applies the same interim decision calculation to an externally observed data cut.

All three routes ultimately use the same interim decision calculation. This vignette separates the package-level flow from the more detailed Bayesian-survival calculation so that the performance paths do not obscure the statistical model.

In the diagrams, **blue** nodes are exported functions, **grey** nodes are general internal calculations, and **gold** nodes are repeated calculations that receive data already prepared and validated by `goldilocks`. The gold and grey routes implement the same statistical analysis.

## Trial simulation and interim evaluation

```{r, echo = FALSE, fig.width = 10, fig.height = 6.2, fig.alt = "Flowchart of simulated trials and external interim evaluations through the shared Goldilocks decision engine."}
DiagrammeR::grViz("
digraph trial_pipeline {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.28, ranksep = 0.48]
  node [shape = box, style = 'filled,rounded', fontname = Helvetica, fontsize = 9, fillcolor = '#f5f5f5', color = '#999999']
  edge [fontname = Helvetica, fontsize = 8, color = '#666666']

  subgraph cluster_entry {
    label = 'Entry points'
    color = '#cccccc'
    style = 'rounded,dashed'
    sim_trials      [label = 'sim_trials()', fillcolor = '#dae8fc', color = '#6c8ebf']
    survival_adapt  [label = 'survival_adapt()', fillcolor = '#dae8fc', color = '#6c8ebf']
    evaluate_interim [label = 'evaluate_interim()', fillcolor = '#dae8fc', color = '#6c8ebf']
  }

  execution_plan [label = 'backend plan + per-trial RNG streams']
  prepare_external [label = 'prepare_external_interim_data()']
  interim_decision [label = 'evaluate_interim_decision()']
  predictive_survival [label = 'analyse_predictive_survival()']
  analyse_final [label = 'analyse_final()']

  subgraph cluster_generation {
    label = 'Trial generation'
    color = '#cccccc'
    style = 'rounded,dashed'
    sim_comp_data [label = 'sim_comp_data()', fillcolor = '#dae8fc', color = '#6c8ebf']
    enrollment [label = 'enrollment()', fillcolor = '#dae8fc', color = '#6c8ebf']
    randomization [label = 'randomization()', fillcolor = '#dae8fc', color = '#6c8ebf']
    pwe_sim [label = 'pwe_sim()', fillcolor = '#dae8fc', color = '#6c8ebf']
  }

  posterior_stats [label = 'posterior_sufficient_stats()']
  posterior_general [label = 'posterior_from_sufficient_stats()']
  posterior_complete [label = 'posterior()']
  impute_batch [label = 'impute_predictive_draws()\nPWE / Bernoulli sampling across imputations']
  prepared_outcomes [label = 'prepare_predictive_outcomes()\nfixed outcomes + imputation positions']
  patient_reference [label = 'complete_predictive_data()\nreference comparisons']
  binary_counts [label = 'predictive_binary_counts()\narm-level event / subject counts']
  impute_data [label = 'impute_data()\nscalar final/reference path']
  pwe_impute [label = 'pwe_impute()', fillcolor = '#dae8fc', color = '#6c8ebf']
  analyse_data [label = 'analyse_data()']
  bayes_prepared [label = 'Prepared Bayesian-survival analysis', fillcolor = '#fff2cc', color = '#d6b656']
  survival_tests [label = 'Completed log-rank / Cox analyses', fillcolor = '#fff2cc', color = '#d6b656']
  binary_analyses [label = 'Binary count analyses\nreuse identical summaries', fillcolor = '#fff2cc', color = '#d6b656']
  rubin_pool [label = 'Cox / risk-difference fits + Rubin pooling']
  prop_to_haz [label = 'prop_to_haz()', fillcolor = '#dae8fc', color = '#6c8ebf']

  sim_trials -> execution_plan
  execution_plan -> survival_adapt [label = 'once per trial']

  survival_adapt -> sim_comp_data
  survival_adapt -> interim_decision [label = 'each interim look']
  survival_adapt -> analyse_final [label = 'stopped or maximum N']

  evaluate_interim -> prepare_external
  prepare_external -> interim_decision

  sim_comp_data -> enrollment
  sim_comp_data -> randomization
  sim_comp_data -> pwe_sim [label = 'generation_cutpoints']
  prop_to_haz -> sim_comp_data [label = 'optional design conversion', style = dashed]

  interim_decision -> posterior_stats [label = 'observed data']
  posterior_stats -> posterior_general [label = 'observed sufficient statistics']
  posterior_general -> impute_batch [label = 'N_impute hazard draws']
  interim_decision -> impute_batch [label = 'one batch per look']
  interim_decision -> prepared_outcomes [label = 'observed outcomes']
  impute_batch -> prepared_outcomes [label = 'survival / regression methods']
  prepared_outcomes -> predictive_survival [label = 'one completed outcome per draw']
  impute_batch -> patient_reference [label = 'reference checks', style = dashed]
  impute_batch -> binary_counts [label = 'fixed-horizon binary methods']
  binary_counts -> binary_analyses [label = 'current + maximum summaries']

  predictive_survival -> bayes_prepared [label = 'bayes-surv']
  predictive_survival -> survival_tests [label = 'logrank / cox']

  analyse_final -> posterior_complete [label = 'if imputed_final']
  posterior_complete -> impute_data [label = 'N_impute hazard draws']
  impute_data -> pwe_impute [label = 'pending follow-up']
  analyse_final -> bayes_prepared [label = 'imputed bayes-surv']
  analyse_final -> analyse_data [label = 'observed / bayes-bin']
  analyse_final -> rubin_pool [label = 'imputed cox / riskdiff']
}
")
```

`sim_trials()` first resolves sequential, fork, or PSOCK execution and, when a seed is supplied, constructs an independent `"L'Ecuyer-CMRG"` stream for every requested trial. Trial-level errors are isolated and retained in the returned failure table. Each successful task calls `survival_adapt()` with the same normalized design but its own random-number stream.

`survival_adapt()` calls `sim_comp_data()` to generate enrollment, treatment assignments, event times, and loss to follow-up. `generation_cutpoints` defines only the event-time data-generating partition. The separate `cutpoints` argument defines posterior estimation, predictive imputation, and Bayesian-survival analysis.

At every simulated look, `survival_adapt()` prepares the observed data cut and calls `evaluate_interim_decision()`. `evaluate_interim()` instead validates an external data cut, derives potential future allocation from `N_total` and `rand_ratio`, and then calls the same decision function. Consequently, simulated and external interim decisions share posterior prediction, imputation, completed-data analysis, threshold comparisons, diagnostics, and trace construction.

The interim calculation draws `N_impute` hazards from the observed-data posterior and passes the complete array to `impute_predictive_draws()`. That function generates subject-by-draw time and event matrices for currently pending follow-up and, when futility is enabled, future subjects through the maximum sample size. It stores only rows requiring imputation. For survival, log-rank, and Cox analyses, the observed outcomes and imputation positions are prepared once per look. Each matrix column is then inserted into copies of the follow-up and event values; fixed treatment assignments are reused. The completed outcomes enter the same log-rank, Cox, or Bayesian-survival calculation used for an ordinary completed dataset. Fixed-horizon binary analyses instead combine each column with the observed event counts and pass the resulting arm-level event and subject totals directly to the same risk-difference or beta-binomial calculations used by the general patient-level analysis. The scalar `impute_data()` helper remains in active use for final-analysis imputation and as the reference implementation, while `complete_predictive_data()` provides a patient-level reference for regression tests.

For deterministic binary analyses, repeated current or maximum-sample count summaries within one interim look are analyzed once and the result is reused. Two summaries match when their control and treatment event counts, sample sizes, and fixed analysis settings all match. Bayesian binary analysis with `bin_method = "mc"` does not reuse results: every predictive completion receives fresh beta-posterior draws in the established current-then-maximum order. The count-based route therefore changes only the representation, not the completed-data model or Monte Carlo procedure. `evaluate_interim()` reports the number of requested, distinct, repeated, and reused count summaries in `diagnostics$binary_count_reuse`.

Within an interim batch, random inputs are generated in posterior-draw order; for each draw the order is current treatment, current control, future treatment, then future control. All predictive imputations are generated before completed-data analysis begins. This remains exactly reproducible under `set.seed()` and across supported serial and parallel backends. Because Bayesian completed-data analyses previously consumed random numbers between imputations, the same seed can produce different Bayesian results than in earlier package versions. The resulting success indicators estimate the predictive probabilities used for expected-success and futility decisions. Exact Monte Carlo bounds are retained as diagnostics; decisions use the point estimates.

## General and prepared Bayesian-survival routes

Bayesian-survival analysis uses the same Gamma piecewise-exponential model on both routes below. The difference is where validation occurs and which intermediate results are constructed.

```{r, echo = FALSE, fig.width = 10, fig.height = 6.2, fig.alt = "Flowchart comparing general and prepared Bayesian-survival analysis routes that share sufficient statistics, Gamma sampling, and posterior summaries."}
DiagrammeR::grViz("
digraph bayes_survival {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.35, ranksep = 0.38]
  node [shape = box, style = 'filled,rounded', fontname = Helvetica, fontsize = 9]
  edge [fontname = Helvetica, fontsize = 8, color = '#666666']

  summary [label = 'summarise_bayes_surv_effect_draws()', fillcolor = '#f5f5f5', color = '#999999']

  subgraph cluster_general {
    label = 'General route'
    color = '#6c8ebf'
    style = 'rounded,dashed'
    analyse_data [label = 'analyse_data()', fillcolor = '#f5f5f5', color = '#999999']
    general_stats [label = 'posterior_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    analyse_general [label = 'analyse_bayes_surv_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    posterior_general [label = 'posterior_from_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    general_gamma [label = 'draw_gamma_posterior()', fillcolor = '#f5f5f5', color = '#999999']
    posterior_draws [label = 'analyse_bayes_surv_posterior_draws()', fillcolor = '#f5f5f5', color = '#999999']
    haz_to_prop [label = 'haz_to_prop()', fillcolor = '#f5f5f5', color = '#999999']
    ppwe [label = 'ppwe()', fillcolor = '#dae8fc', color = '#6c8ebf']

    analyse_data -> general_stats
    general_stats -> analyse_general [label = 'validated and normalized']
    analyse_general -> posterior_general
    posterior_general -> general_gamma
    general_gamma -> posterior_draws
    posterior_draws -> haz_to_prop
    haz_to_prop -> ppwe [label = 'piecewise hazards']
  }

  subgraph cluster_prepared {
    label = 'Prepared repeated route'
    color = '#d6b656'
    style = 'rounded,dashed'
    prepared_callers [label = 'analyse_predictive_survival() / analyse_final()', fillcolor = '#f5f5f5', color = '#999999']
    prepared_stats [label = 'summarise_survival_outcomes()\nor general data-frame function', fillcolor = '#f5f5f5', color = '#999999']
    analyse_prepared [label = 'analyse_prepared_bayes_surv()', fillcolor = '#fff2cc', color = '#d6b656']
    posterior_prepared [label = 'posterior_from_prepared_stats()', fillcolor = '#fff2cc', color = '#d6b656']
    prepared_gamma [label = 'draw_gamma_posterior()', fillcolor = '#f5f5f5', color = '#999999']
    effect_prepared [label = 'bayes_surv_effect_draws()', fillcolor = '#fff2cc', color = '#d6b656']
    interval_widths [label = 'endpoint_interval_widths()', fillcolor = '#fff2cc', color = '#d6b656']

    prepared_callers -> prepared_stats [label = 'completed outcome vectors / data']
    prepared_stats -> analyse_prepared
    analyse_prepared -> posterior_prepared
    posterior_prepared -> prepared_gamma
    prepared_gamma -> effect_prepared
    interval_widths -> effect_prepared [label = 'fixed horizon']
  }

  haz_to_prop -> summary [label = 'effect draws']
  effect_prepared -> summary [label = 'effect draws']
}
")
```

The general route accepts the package's supported prior specifications and verifies sufficient-statistic columns, arm-interval combinations, counts, and empty intervals while normalizing arbitrary valid row order. It remains the route used by ordinary `analyse_data()` calls, including observed-data Bayesian-survival final analyses. `haz_to_prop()` returns treatment, control, and effect draws and uses the exported `ppwe()` calculation for piecewise hazards.

The prepared route is used only after `goldilocks` has produced validated completed outcomes and normalized the Gamma priors. At interim looks, `summarise_survival_outcomes()` calculates arm-by-interval events and exposure directly from the completed outcomes. Imputed final analyses use the general data-frame function to produce the same sufficient-statistic layout. Both then retain inexpensive consistency checks, draw from exactly the same Gamma posterior, and consume random numbers in exactly the same order as the general route. The prepared calculation returns only the treatment-effect draws needed by the decision rule, avoiding repeated validation and temporary probability data frames.

For analysis cutpoints $0 < c_1 < \cdots < c_{J-1} < \tau$, where $\tau$ is `end_of_study`, `endpoint_interval_widths()` returns

\[
\mathbf{w} = (c_1,\ c_2-c_1,\ \ldots,\ \tau-c_{J-1}).
\]

For arm \(a\), a posterior hazard draw is converted to cumulative hazard and event probability using

\[
H_a(\tau) = \sum_{j=1}^{J}\lambda_{aj}w_j,
\qquad
p_a(\tau) = 1 - \exp\{-H_a(\tau)\}.
\]

These widths describe the fixed analysis horizon, not the maximum follow-up currently observed. Observed events and person-time enter separately through the sufficient statistics used in the Gamma posterior. If nobody has entered a late interval, `empty_interval = "prior"` leaves that interval prior-driven; `"propagate"` applies the configured adjacent-interval rule; and `"error"` stops. Shortening \(\mathbf{w}\) to observed follow-up would change the estimand rather than correct the cumulative-hazard calculation.

## Completed-data analysis methods

`analyse_data()` applies the selected `method` to a completed dataset. Interim prediction analyzes each completed dataset independently before averaging its binary success indicators. For fixed-horizon binary methods, event counts are sufficient for that analysis: the patient-level and count-based predictive routes use the same count calculations.

| `method` | Completed-data analysis | `imputed_final = TRUE` | `imputed_final = FALSE` |
|:--|:--|:--|:--|
| `bayes-surv` | Gamma piecewise-exponential posterior; effect is the treatment-minus-control event probability, or the treatment probability in a single-arm design | Analyze each completed imputation through the prepared sufficient-statistics and effect calculations, then average posterior summaries | Analyze observed right-censored data through the general route |
| `bayes-bin` | Beta-binomial posterior for completed binary endpoint status | Analyze each completed imputation and average posterior summaries | Exclude subjects lost before complete endpoint ascertainment |
| `cox` | Cox Wald test for the log hazard ratio | Pool estimates and within-imputation variances using Rubin's rules | Analyze observed right-censored data |
| `riskdiff` | Wald test for the treatment-minus-control event-risk difference | Pool estimates and within-imputation variances using Rubin's rules | Exclude subjects lost before complete endpoint ascertainment |
| `logrank` | Log-rank test | Not supported because no imputation-pooling rule is implemented | Analyze observed right-censored data |

The posterior-imputation model and completed-data analysis model are deliberately modular. In particular, `method = "bayes-bin"` still uses the piecewise-exponential Gamma model to impute pending outcomes, then applies the separate beta-binomial model to completed endpoint statuses.

## Results and post-processing

```{r, echo = FALSE, fig.width = 10, fig.height = 4.2, fig.alt = "Flowchart from Goldilocks entry-point results to summaries and plots."}
DiagrammeR::grViz("
digraph results {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.25, ranksep = 0.48]
  node [shape = box, style = 'filled,rounded', fontname = Helvetica, fontsize = 9]
  edge [fontname = Helvetica, fontsize = 8, color = '#666666']

  survival_adapt [label = 'survival_adapt()', fillcolor = '#dae8fc', color = '#6c8ebf']
  evaluate_interim [label = 'evaluate_interim()', fillcolor = '#dae8fc', color = '#6c8ebf']
  sim_trials [label = 'sim_trials()', fillcolor = '#dae8fc', color = '#6c8ebf']

  trial_result [label = 'trial summary + optional trace\nretained design and diagnostics', fillcolor = '#f5f5f5', color = '#999999']
  interim_result [label = 'decision + predictive probabilities\ndiagnostics + one-look trace', fillcolor = '#f5f5f5', color = '#999999']
  sim_result [label = 'trial summaries + failures + optional traces\nRNG, parallel, timing, and design metadata', fillcolor = '#f5f5f5', color = '#999999']
  oc_summary [label = 'summarise_sims()', fillcolor = '#dae8fc', color = '#6c8ebf']

  trace_outputs [label = 'summarise_trial_trace()\nplot_trial_trace()', fillcolor = '#dae8fc', color = '#6c8ebf']
  enrollment_plot [label = 'plot_enrollment()', fillcolor = '#dae8fc', color = '#6c8ebf']
  calendar_summary [label = 'summarise_calendar_time()', fillcolor = '#dae8fc', color = '#6c8ebf']
  sim_plots [label = 'plot_sim_stopping()\nplot_sim_decisions()', fillcolor = '#dae8fc', color = '#6c8ebf']
  oc_plot [label = 'plot_sim_ocs()', fillcolor = '#dae8fc', color = '#6c8ebf']

  survival_adapt -> trial_result
  evaluate_interim -> interim_result
  sim_trials -> sim_result

  trial_result -> trace_outputs
  interim_result -> trace_outputs
  trial_result -> enrollment_plot
  sim_result -> enrollment_plot
  sim_result -> calendar_summary
  sim_result -> sim_plots
  sim_result -> oc_summary
  oc_summary -> oc_plot
}
")
```

## Exported utilities and retained outputs

The exported low-level utilities are:

-   `enrollment()` for exact continuous-time accrual;
-   `randomization()` for blocked two-arm allocation;
-   `pwe_sim()` and `pwe_impute()` for unconditional and conditional piecewise-exponential event times;
-   `ppwe()` for piecewise-exponential cumulative event probabilities;
-   `prop_to_haz()` for converting cumulative event probabilities to piecewise hazards; and
-   `sim_comp_data()` for generating a complete simulated trial dataset.

The three main entry points retain enough information for audit and downstream displays:

-   `survival_adapt()` returns the one-row trial summary and optionally a compact interim trace, with normalized arguments, decision design, prior diagnostics, enrollment design, and calendar metrics attached;
-   `sim_trials()` returns successful trial summaries, isolated failure records, optional traces, RNG and parallel metadata, and the common retained design; and
-   `evaluate_interim()` returns the decision, predictive probabilities, Monte Carlo diagnostics, allocation and data-cut diagnostics, posterior diagnostics, binary count-summary reuse diagnostics when applicable, and a compatible one-look trace.

Post-processing functions consume these retained results without rerunning the trial:

-   `summarise_sims()` calculates operating characteristics and Monte Carlo uncertainty;
-   `summarise_calendar_time()` summarizes trial duration, accrual, analysis readiness, follow-up, and interim timing;
-   `summarise_trial_trace()` condenses a single trial's stopping path;
-   `plot_enrollment()` displays expected or retained accrual and analysis milestones;
-   `plot_trial_trace()` displays one trial's predictive-probability path;
-   `plot_sim_stopping()` summarizes stopping outcomes and sample sizes;
-   `plot_sim_decisions()` displays interim decision regions across simulated trials; and
-   `plot_sim_ocs()` compares operating characteristics across scenarios.

## Architectural invariants

Several conventions link the layers above:

-   data use `treatment = 0` for control and `treatment = 1` for treatment;
-   normalized arm-specific vectors and priors use canonical `control`, `treatment` order, while posterior hazard arrays retain treatment in slice 1 and control in slice 2 for compatibility;
-   `generation_cutpoints` governs event-time generation only, whereas `cutpoints` governs posterior estimation, predictive imputation, analysis priors, and endpoint probability calculations;
-   analysis intervals follow the survival counting-process convention `(start, stop]` when observed events are assigned to sufficient statistics;
-   outer predictive decisions use `N_impute`, completed-data Bayesian Monte Carlo uses `N_mcmc`, and exact bounds are diagnostic rather than decision-changing; and
-   interim RNG order is posterior hazards, the complete predictive-imputation batch in draw/current/future/arm order, and then completed-data analyses; Bayesian binary Monte Carlo retains current-then-maximum order within each draw and does not reuse repeated summaries; and
-   prepared repeated calculations may bypass rich-input validation only after package-owned functions have constructed canonical inputs. General entry points retain full validation.
