---
title: "Package architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `goldilocks` package implements the Goldilocks adaptive trial design described in Broglio et al. (2014). This vignette provides a visual overview of how the package functions are interconnected.

## Function dependency diagram

The diagram below shows the call graph from the top-level simulation function (`sim_trials()`) down through the core engine (`survival_adapt()`) and into the internal analysis pipeline.

**Exported functions** are shown in blue. **Internal functions** are shown in grey.

```{r, echo = FALSE, fig.width = 8, fig.height = 8, fig.alt = "Diagram showing the function call dependencies in the goldilocks package."}
DiagrammeR::grViz("
digraph goldilocks {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.4, ranksep = 0.6]

  node [shape = box, style = 'filled, rounded', fontname = Helvetica, fontsize = 10]

  # Exported functions (blue)
  sim_trials       [label = 'sim_trials()',       fillcolor = '#dae8fc', color = '#6c8ebf']
  survival_adapt   [label = 'survival_adapt()',   fillcolor = '#dae8fc', color = '#6c8ebf']
  summarise_sims   [label = 'summarise_sims()',   fillcolor = '#dae8fc', color = '#6c8ebf']
  summarise_trial_trace [label = 'summarise_trial_trace()', fillcolor = '#dae8fc', color = '#6c8ebf']
  plot_trial_trace [label = 'plot_trial_trace()', fillcolor = '#dae8fc', color = '#6c8ebf']
  plot_sim_stopping [label = 'plot_sim_stopping()', fillcolor = '#dae8fc', color = '#6c8ebf']
  plot_sim_ocs      [label = 'plot_sim_ocs()', fillcolor = '#dae8fc', color = '#6c8ebf']
  plot_sim_decisions [label = 'plot_sim_decisions()', fillcolor = '#dae8fc', color = '#6c8ebf']
  sim_comp_data    [label = 'sim_comp_data()',    fillcolor = '#dae8fc', color = '#6c8ebf']
  enrollment       [label = 'enrollment()',       fillcolor = '#dae8fc', color = '#6c8ebf']
  randomization    [label = 'randomization()',    fillcolor = '#dae8fc', color = '#6c8ebf']
  pwe_sim          [label = 'pwe_sim()',          fillcolor = '#dae8fc', color = '#6c8ebf']
  pwe_impute       [label = 'pwe_impute()',       fillcolor = '#dae8fc', color = '#6c8ebf']
  ppwe             [label = 'ppwe()',             fillcolor = '#dae8fc', color = '#6c8ebf']

  # Internal functions (grey)
  test_stop_success [label = 'test_stop_success()', fillcolor = '#f5f5f5', color = '#999999']
  test_final        [label = 'test_final()',         fillcolor = '#f5f5f5', color = '#999999']
  analyse_data      [label = 'analyse_data()',       fillcolor = '#f5f5f5', color = '#999999']
  bayes_binomial_test [label = 'bayes_binomial_test()', fillcolor = '#f5f5f5', color = '#999999']
  impute_data       [label = 'impute_data()',        fillcolor = '#f5f5f5', color = '#999999']
  posterior          [label = 'posterior()',           fillcolor = '#f5f5f5', color = '#999999']
  haz_to_prop       [label = 'haz_to_prop()',        fillcolor = '#f5f5f5', color = '#999999']
  logrank_test      [label = 'logrank_test()',        fillcolor = '#f5f5f5', color = '#999999']

  # Edges
  sim_trials      -> survival_adapt
  sim_trials      -> summarise_sims  [style = dashed, label = 'output list']
  sim_trials      -> plot_sim_stopping [style = dashed, label = 'simulation output']
  sim_trials      -> plot_sim_decisions [style = dashed, label = 'simulation traces']
  summarise_sims  -> plot_sim_ocs [style = dashed, label = 'scenario summaries']

  survival_adapt  -> sim_comp_data
  survival_adapt  -> posterior
  survival_adapt  -> test_stop_success
  survival_adapt  -> test_final
  survival_adapt  -> summarise_trial_trace [style = dashed, label = 'optional trace']
  survival_adapt  -> plot_trial_trace [style = dashed, label = 'optional trace']

  sim_comp_data   -> enrollment
  sim_comp_data   -> randomization
  sim_comp_data   -> pwe_sim

  test_stop_success -> impute_data
  test_stop_success -> analyse_data

  test_final      -> posterior
  test_final      -> impute_data
  test_final      -> analyse_data

  analyse_data    -> posterior
  analyse_data    -> haz_to_prop
  analyse_data    -> bayes_binomial_test
  analyse_data    -> logrank_test

  haz_to_prop     -> ppwe

  impute_data     -> pwe_impute
  impute_data     -> pwe_sim
}
")
```

## Function roles

The functions fall into three layers:

### Simulation layer

-   **`sim_comp_data()`**: Generates a complete trial dataset by calling `enrollment()`, `randomization()`, and `pwe_sim()`.
-   **`survival_adapt()`**: Simulates a single adaptive trial. Generates data via `sim_comp_data()`, conducts interim analyses using `posterior()` and `test_stop_success()`, and performs the final analysis via `test_final()`. With `return_trace = TRUE`, it also retains a compact audit trail for each completed interim look.
-   **`sim_trials()`**: Top-level entry point. Runs `survival_adapt()` across multiple trials (optionally in parallel) and collates results.

### Post-processing functions

-   **`summarise_sims()`**: Summarizes the output of `sim_trials()`, computing operating characteristics such as power, expected sample size, and stopping probabilities.
-   **`summarise_trial_trace()`**: Condenses an optional single-trial interim trace into a one-row stopping-path summary.
-   **`plot_trial_trace()`**: Visualizes predictive probabilities, thresholds, enrollment, and observed events for an optional single-trial trace.
-   **`plot_sim_stopping()`**: Visualizes marginal, conditional, cumulative, or flowchart stopping outcomes and enrolled sample sizes across simulated trials.
-   **`plot_sim_ocs()`**: Compares success, stopping, and expected-sample-size operating characteristics across treatment-effect scenarios.
-   **`plot_sim_decisions()`**: Maps simulated predictive probabilities into expected-success, continuation, and futility regions at each interim look.

### Data generation and analysis utilities

-   **`posterior()`**: Estimates the posterior distribution of piecewise exponential hazard rates using a conjugate Gamma model.
-   **`analyse_data()`**: Applies the chosen analysis method (`logrank`, `cox`, `bayes-surv`, `bayes-bin`, or `riskdiff`) to an (imputed) dataset.
-   **`impute_data()`**: Imputes missing event times for censored subjects using `pwe_impute()` or `pwe_sim()`.
-   **`haz_to_prop()`**: Converts posterior hazard rate draws to cumulative incidence proportions via `ppwe()`.
