---
title: "Package architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `goldilocks` package has three ways into the adaptive calculation:

1.  `sim_trials()` simulates many independent trials and delegates each one to `survival_adapt()`;
2.  `survival_adapt()` generates and evaluates one complete simulated trial; and
3.  `evaluate_interim()` applies the same interim decision calculation to an externally observed data cut.

All three routes ultimately use the same interim decision engine. This vignette separates the package-level flow from the more detailed Bayesian-survival calculation so that the performance paths do not obscure the statistical model.

In the diagrams, **blue** nodes are exported functions, **grey** nodes are internal orchestration or checked analysis functions, and **gold** nodes are trusted internal kernels. A trusted kernel receives canonical objects that have already been validated and normalized by `goldilocks`; it does not define a different statistical analysis.

## Trial simulation and interim evaluation

```{r, echo = FALSE, fig.width = 10, fig.height = 6.2, fig.alt = "Flowchart of simulated trials and external interim evaluations through the shared Goldilocks decision engine."}
DiagrammeR::grViz("
digraph trial_pipeline {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.28, ranksep = 0.48]
  node [shape = box, style = 'filled,rounded', fontname = Helvetica, fontsize = 9, fillcolor = '#f5f5f5', color = '#999999']
  edge [fontname = Helvetica, fontsize = 8, color = '#666666']

  subgraph cluster_entry {
    label = 'Entry points'
    color = '#cccccc'
    style = 'rounded,dashed'
    sim_trials      [label = 'sim_trials()', fillcolor = '#dae8fc', color = '#6c8ebf']
    survival_adapt  [label = 'survival_adapt()', fillcolor = '#dae8fc', color = '#6c8ebf']
    evaluate_interim [label = 'evaluate_interim()', fillcolor = '#dae8fc', color = '#6c8ebf']
  }

  execution_plan [label = 'backend plan + per-trial RNG streams']
  prepare_external [label = 'prepare_external_interim_data()']
  interim_core [label = 'evaluate_interim_core()']
  test_stop [label = 'test_stop_success()']
  test_final [label = 'test_final()']

  subgraph cluster_generation {
    label = 'Trial generation'
    color = '#cccccc'
    style = 'rounded,dashed'
    sim_comp_data [label = 'sim_comp_data()', fillcolor = '#dae8fc', color = '#6c8ebf']
    enrollment [label = 'enrollment()', fillcolor = '#dae8fc', color = '#6c8ebf']
    randomization [label = 'randomization()', fillcolor = '#dae8fc', color = '#6c8ebf']
    pwe_sim [label = 'pwe_sim()', fillcolor = '#dae8fc', color = '#6c8ebf']
  }

  posterior_stats [label = 'posterior_sufficient_stats()']
  posterior_checked [label = 'posterior_from_sufficient_stats()']
  posterior_complete [label = 'posterior()']
  impute_batch [label = 'impute_predictive_draws()\nvectorized PWE / Bernoulli sampling']
  materialize [label = 'materialize_predictive_draw()']
  impute_data [label = 'impute_data()\nscalar final/reference path']
  pwe_impute [label = 'pwe_impute()', fillcolor = '#dae8fc', color = '#6c8ebf']
  analyse_data [label = 'analyse_data()']
  bayes_kernel [label = 'Bayesian-survival kernels', fillcolor = '#fff2cc', color = '#d6b656']
  rubin_pool [label = 'Cox / risk-difference fits + Rubin pooling']
  prop_to_haz [label = 'prop_to_haz()', fillcolor = '#dae8fc', color = '#6c8ebf']

  sim_trials -> execution_plan
  execution_plan -> survival_adapt [label = 'once per trial']

  survival_adapt -> sim_comp_data
  survival_adapt -> interim_core [label = 'each interim look']
  survival_adapt -> test_final [label = 'stopped or maximum N']

  evaluate_interim -> prepare_external
  prepare_external -> interim_core

  sim_comp_data -> enrollment
  sim_comp_data -> randomization
  sim_comp_data -> pwe_sim [label = 'generation_cutpoints']
  prop_to_haz -> sim_comp_data [label = 'optional design conversion', style = dashed]

  interim_core -> posterior_stats [label = 'observed data']
  posterior_stats -> posterior_checked [label = 'observed sufficient statistics']
  posterior_checked -> impute_batch [label = 'N_impute hazard draws']
  interim_core -> impute_batch [label = 'one batch per look']
  impute_batch -> materialize [label = 'time / event matrices']
  materialize -> test_stop [label = 'one column per analysis']

  test_stop -> bayes_kernel [label = 'bayes-surv']
  test_stop -> analyse_data [label = 'other methods']

  test_final -> posterior_complete [label = 'if imputed_final']
  posterior_complete -> impute_data [label = 'N_impute hazard draws']
  impute_data -> pwe_impute [label = 'pending follow-up']
  test_final -> bayes_kernel [label = 'imputed bayes-surv']
  test_final -> analyse_data [label = 'observed / bayes-bin']
  test_final -> rubin_pool [label = 'imputed cox / riskdiff']
}
")
```

`sim_trials()` first resolves sequential, fork, or PSOCK execution and, when a seed is supplied, constructs an independent `"L'Ecuyer-CMRG"` stream for every requested trial. Trial-level errors are isolated and retained in the returned failure table. Each successful task calls `survival_adapt()` with the same normalized design but its own random-number stream.

`survival_adapt()` calls `sim_comp_data()` to generate enrollment, treatment assignments, event times, and loss to follow-up. `generation_cutpoints` defines only the event-time data-generating partition. The separate `cutpoints` argument defines posterior estimation, predictive imputation, and Bayesian-survival analysis.

At every simulated look, `survival_adapt()` prepares the observed data cut and calls `evaluate_interim_core()`. `evaluate_interim()` instead validates an external data cut, derives potential future allocation from `N_total` and `rand_ratio`, and then calls the same core function. Consequently, simulated and external interim decisions share posterior prediction, imputation, completed-data analysis, threshold comparisons, diagnostics, and trace construction.

The core draws `N_impute` hazards from the observed-data posterior and passes the complete array to `impute_predictive_draws()`. That internal batch calculation generates subject-by-draw time and event matrices for currently pending follow-up and, when futility is enabled, future subjects through the maximum sample size. It stores only rows requiring imputation. `test_stop_success()` then materializes and analyzes one matrix column at a time, so the completed-data analysis models and decision rules remain unchanged. The scalar `impute_data()` helper remains in active use for final-analysis imputation and as the reference implementation.

Within an interim batch, random inputs are generated in posterior-draw order; for each draw the order is current treatment, current control, future treatment, then future control. All predictive imputations are generated before completed-data analysis begins. This remains exactly reproducible under `set.seed()` and across supported serial and parallel backends. Because Bayesian completed-data analyses previously consumed random numbers between imputations, the same seed can produce different Bayesian results than in earlier package versions. The resulting success indicators estimate the predictive probabilities used for expected-success and futility decisions. Exact Monte Carlo bounds are retained as diagnostics; decisions use the point estimates.

## Bayesian-survival checked and trusted paths

Bayesian-survival analysis uses the same Gamma piecewise-exponential model on both paths below. The difference is where validation occurs and which intermediate objects are materialized.

```{r, echo = FALSE, fig.width = 10, fig.height = 6.2, fig.alt = "Flowchart comparing checked and trusted Bayesian-survival analysis paths that share sufficient statistics, Gamma sampling, and posterior summaries."}
DiagrammeR::grViz("
digraph bayes_survival {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.35, ranksep = 0.38]
  node [shape = box, style = 'filled,rounded', fontname = Helvetica, fontsize = 9]
  edge [fontname = Helvetica, fontsize = 8, color = '#666666']

  summary [label = 'summarise_bayes_surv_effect_draws()', fillcolor = '#f5f5f5', color = '#999999']

  subgraph cluster_checked {
    label = 'General checked path'
    color = '#6c8ebf'
    style = 'rounded,dashed'
    analyse_data [label = 'analyse_data()', fillcolor = '#f5f5f5', color = '#999999']
    checked_stats [label = 'posterior_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    analyse_checked [label = 'analyse_bayes_surv_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    posterior_checked [label = 'posterior_from_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    checked_gamma [label = 'draw_gamma_posterior_kernel()', fillcolor = '#f5f5f5', color = '#999999']
    posterior_draws [label = 'analyse_bayes_surv_posterior_draws()', fillcolor = '#f5f5f5', color = '#999999']
    haz_to_prop [label = 'haz_to_prop()', fillcolor = '#f5f5f5', color = '#999999']
    ppwe [label = 'ppwe()', fillcolor = '#dae8fc', color = '#6c8ebf']

    analyse_data -> checked_stats
    checked_stats -> analyse_checked [label = 'validated and normalized']
    analyse_checked -> posterior_checked
    posterior_checked -> checked_gamma
    checked_gamma -> posterior_draws
    posterior_draws -> haz_to_prop
    haz_to_prop -> ppwe [label = 'piecewise hazards']
  }

  subgraph cluster_trusted {
    label = 'Repeated trusted path'
    color = '#d6b656'
    style = 'rounded,dashed'
    trusted_callers [label = 'test_stop_success() / test_final()', fillcolor = '#f5f5f5', color = '#999999']
    trusted_stats [label = 'posterior_sufficient_stats()', fillcolor = '#f5f5f5', color = '#999999']
    analyse_kernel [label = 'analyse_bayes_surv_sufficient_stats_kernel()', fillcolor = '#fff2cc', color = '#d6b656']
    posterior_kernel [label = 'posterior_from_sufficient_stats_kernel()', fillcolor = '#fff2cc', color = '#d6b656']
    trusted_gamma [label = 'draw_gamma_posterior_kernel()', fillcolor = '#f5f5f5', color = '#999999']
    effect_kernel [label = 'bayes_surv_effect_draws_kernel()', fillcolor = '#fff2cc', color = '#d6b656']
    interval_widths [label = 'endpoint_interval_widths()', fillcolor = '#fff2cc', color = '#d6b656']

    trusted_callers -> trusted_stats [label = 'canonical completed data']
    trusted_stats -> analyse_kernel
    analyse_kernel -> posterior_kernel
    posterior_kernel -> trusted_gamma
    trusted_gamma -> effect_kernel
    interval_widths -> effect_kernel [label = 'fixed horizon']
  }

  haz_to_prop -> summary [label = 'effect draws']
  effect_kernel -> summary [label = 'effect draws']
}
")
```

The checked route accepts the package's supported prior specifications and defensively verifies sufficient-statistic columns, arm-interval combinations, counts, and empty intervals while normalizing arbitrary valid row order. It remains the route used by ordinary `analyse_data()` calls, including observed-data Bayesian-survival final analyses. `haz_to_prop()` returns treatment, control, and effect draws and uses the exported `ppwe()` calculation for piecewise hazards.

The trusted route is used only after `goldilocks` has produced canonical arm-by-interval sufficient statistics and normalized the Gamma priors. It retains inexpensive invariant checks, draws from exactly the same Gamma posterior, and consumes random numbers in exactly the same order as the checked route. It then calculates only the treatment-effect draws needed by the decision rule, avoiding repeated cutpoint validation and temporary probability data frames.

For analysis cutpoints $0 < c_1 < \cdots < c_{J-1} < \tau$, where $\tau$ is `end_of_study`, `endpoint_interval_widths()` returns

\[
\mathbf{w} = (c_1,\ c_2-c_1,\ \ldots,\ \tau-c_{J-1}).
\]

For arm \(a\), a posterior hazard draw is converted to cumulative hazard and event probability using

\[
H_a(\tau) = \sum_{j=1}^{J}\lambda_{aj}w_j,
\qquad
p_a(\tau) = 1 - \exp\{-H_a(\tau)\}.
\]

These widths describe the fixed analysis horizon, not the maximum follow-up currently observed. Observed events and person-time enter separately through the sufficient statistics used in the Gamma posterior. If nobody has entered a late interval, `empty_interval = "prior"` leaves that interval prior-driven; `"propagate"` applies the configured adjacent-interval rule; and `"error"` stops. Shortening \(\mathbf{w}\) to observed follow-up would change the estimand rather than correct the cumulative-hazard calculation.

## Completed-data analysis methods

`analyse_data()` dispatches a completed dataset according to `method`. Interim prediction tests each completed dataset independently before averaging its binary success indicators.

| `method` | Completed-data analysis | `imputed_final = TRUE` | `imputed_final = FALSE` |
|:--|:--|:--|:--|
| `bayes-surv` | Gamma piecewise-exponential posterior; effect is the treatment-minus-control event probability, or the treatment probability in a single-arm design | Analyze each completed imputation through the trusted sufficient-statistics and effect kernels, then average posterior summaries | Analyze observed right-censored data through the checked path |
| `bayes-bin` | Beta-binomial posterior for completed binary endpoint status | Analyze each completed imputation and average posterior summaries | Exclude subjects lost before complete endpoint ascertainment |
| `cox` | Cox Wald test for the log hazard ratio | Pool estimates and within-imputation variances using Rubin's rules | Analyze observed right-censored data |
| `riskdiff` | Wald test for the treatment-minus-control event-risk difference | Pool estimates and within-imputation variances using Rubin's rules | Exclude subjects lost before complete endpoint ascertainment |
| `logrank` | Log-rank test | Not supported because no imputation-pooling rule is implemented | Analyze observed right-censored data |

The posterior-imputation model and completed-data analysis model are deliberately modular. In particular, `method = "bayes-bin"` still uses the piecewise-exponential Gamma model to impute pending outcomes, then applies the separate beta-binomial model to completed endpoint statuses.

## Results and post-processing

```{r, echo = FALSE, fig.width = 10, fig.height = 4.2, fig.alt = "Flowchart from Goldilocks entry-point results to summaries and plots."}
DiagrammeR::grViz("
digraph results {

  graph [rankdir = TB, fontsize = 12, nodesep = 0.25, ranksep = 0.48]
  node [shape = box, style = 'filled,rounded', fontname = Helvetica, fontsize = 9]
  edge [fontname = Helvetica, fontsize = 8, color = '#666666']

  survival_adapt [label = 'survival_adapt()', fillcolor = '#dae8fc', color = '#6c8ebf']
  evaluate_interim [label = 'evaluate_interim()', fillcolor = '#dae8fc', color = '#6c8ebf']
  sim_trials [label = 'sim_trials()', fillcolor = '#dae8fc', color = '#6c8ebf']

  trial_result [label = 'trial summary + optional trace\nretained design and diagnostics', fillcolor = '#f5f5f5', color = '#999999']
  interim_result [label = 'decision + predictive probabilities\ndiagnostics + one-look trace', fillcolor = '#f5f5f5', color = '#999999']
  sim_result [label = 'trial summaries + failures + optional traces\nRNG, parallel, timing, and design metadata', fillcolor = '#f5f5f5', color = '#999999']
  oc_summary [label = 'summarise_sims()', fillcolor = '#dae8fc', color = '#6c8ebf']

  trace_outputs [label = 'summarise_trial_trace()\nplot_trial_trace()', fillcolor = '#dae8fc', color = '#6c8ebf']
  enrollment_plot [label = 'plot_enrollment()', fillcolor = '#dae8fc', color = '#6c8ebf']
  calendar_summary [label = 'summarise_calendar_time()', fillcolor = '#dae8fc', color = '#6c8ebf']
  sim_plots [label = 'plot_sim_stopping()\nplot_sim_decisions()', fillcolor = '#dae8fc', color = '#6c8ebf']
  oc_plot [label = 'plot_sim_ocs()', fillcolor = '#dae8fc', color = '#6c8ebf']

  survival_adapt -> trial_result
  evaluate_interim -> interim_result
  sim_trials -> sim_result

  trial_result -> trace_outputs
  interim_result -> trace_outputs
  trial_result -> enrollment_plot
  sim_result -> enrollment_plot
  sim_result -> calendar_summary
  sim_result -> sim_plots
  sim_result -> oc_summary
  oc_summary -> oc_plot
}
")
```

## Exported utilities and retained outputs

The exported low-level utilities are:

-   `enrollment()` for exact continuous-time accrual;
-   `randomization()` for blocked two-arm allocation;
-   `pwe_sim()` and `pwe_impute()` for unconditional and conditional piecewise-exponential event times;
-   `ppwe()` for piecewise-exponential cumulative event probabilities;
-   `prop_to_haz()` for converting cumulative event probabilities to piecewise hazards; and
-   `sim_comp_data()` for generating a complete simulated trial dataset.

The three main entry points retain enough information for audit and downstream displays:

-   `survival_adapt()` returns the one-row trial summary and optionally a compact interim trace, with normalized arguments, decision design, prior diagnostics, enrollment design, and calendar metrics attached;
-   `sim_trials()` returns successful trial summaries, isolated failure records, optional traces, RNG and parallel metadata, and the common retained design; and
-   `evaluate_interim()` returns the decision, predictive probabilities, Monte Carlo diagnostics, allocation and data-cut diagnostics, posterior diagnostics, and a compatible one-look trace.

Post-processing functions consume these retained results without rerunning the trial:

-   `summarise_sims()` calculates operating characteristics and Monte Carlo uncertainty;
-   `summarise_calendar_time()` summarizes trial duration, accrual, analysis readiness, follow-up, and interim timing;
-   `summarise_trial_trace()` condenses a single trial's stopping path;
-   `plot_enrollment()` displays expected or retained accrual and analysis milestones;
-   `plot_trial_trace()` displays one trial's predictive-probability path;
-   `plot_sim_stopping()` summarizes stopping outcomes and sample sizes;
-   `plot_sim_decisions()` displays interim decision regions across simulated trials; and
-   `plot_sim_ocs()` compares operating characteristics across scenarios.

## Architectural invariants

Several conventions link the layers above:

-   data use `treatment = 0` for control and `treatment = 1` for treatment;
-   normalized arm-specific vectors and priors use canonical `control`, `treatment` order, while posterior hazard arrays retain treatment in slice 1 and control in slice 2 for compatibility;
-   `generation_cutpoints` governs event-time generation only, whereas `cutpoints` governs posterior estimation, predictive imputation, analysis priors, and endpoint probability calculations;
-   analysis intervals follow the survival counting-process convention `(start, stop]` when observed events are assigned to sufficient statistics;
-   outer predictive decisions use `N_impute`, completed-data Bayesian Monte Carlo uses `N_mcmc`, and exact bounds are diagnostic rather than decision-changing; and
-   interim RNG order is posterior hazards, the complete predictive-imputation batch in draw/current/future/arm order, and then completed-data analyses; and
-   trusted kernels may bypass repeated rich-input validation only after package-owned callers have constructed canonical inputs. General entry points retain checked behavior.
