---
title: "Bayesian binary outcome designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian binary outcome designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(5107)
```

```{r setup, message=FALSE}
library(goldilocks)
```

`goldilocks` can analyze completed binary endpoints using
`method = "bayes-bin"`. This is useful when the final decision is based on the
event indicator by a fixed endpoint time, rather than the event time itself.
The trial simulator still uses time-to-event hazards to generate when events
occur and to impute not-yet-observed outcomes at interim looks, but the final
analysis reduces each completed dataset to binary event status by
`end_of_study`.

Two practical consequences follow:

- `method = "bayes-bin"` only supports one-sided alternatives:
  `alternative = "less"` or `alternative = "greater"`.
- The final binary analysis requires complete endpoint status. Subjects who are
  censored before `end_of_study` must be imputed (`imputed_final = TRUE`) or,
  if `imputed_final = FALSE`, excluded from the final binary analysis.

## Two-arm design

Suppose the control event probability by 12 months is 35%, and the treatment is
expected to reduce this to 25%. We use a beta-binomial final analysis with a
uniform `Beta(1, 1)` prior in each arm. Since lower event probability is
beneficial, we use `alternative = "less"` and compare the posterior distribution
of

$$p_{\text{treatment}} - p_{\text{control}}$$

against `h0 = 0`.

```{r two_arm_design}
end_of_study <- 12
hc <- prop_to_haz(0.35, endtime = end_of_study)
ht <- prop_to_haz(0.25, endtime = end_of_study)

out_two_arm <- survival_adapt(
  hazard_treatment = ht,
  hazard_control = hc,
  cutpoints = 0,
  N_total = 120,
  lambda = 10,
  lambda_time = 0,
  interim_look = 80,
  end_of_study = end_of_study,
  bin_prior = c(1, 1),
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0,
  alternative = "less",
  h0 = 0,
  Fn = 0.05,
  Sn = 0.90,
  prob_ha = 0.95,
  N_impute = 20,
  method = "bayes-bin",
  imputed_final = FALSE
)

out_two_arm
```

The output has the same structure as other `survival_adapt()` analyses.
For `method = "bayes-bin"`, `est_final` is the posterior mean binary effect:
the treatment event probability minus the control event probability. The
`post_prob_ha` value is the posterior probability that this difference is below
`h0` when `alternative = "less"`.

## Single-arm design

For a single-arm design, set `hazard_control = NULL`. The comparator is an
external benchmark event probability supplied through `h0`, often called a
performance goal (PG) or objective performance criterion (OPC).

Here the benchmark event probability is 30%, and the target event probability
is 20%. With `alternative = "less"`, success means the posterior probability
that the event probability is below 30% exceeds `prob_ha`.

```{r single_arm_design}
benchmark <- 0.30
target <- 0.20
ht_single <- prop_to_haz(target, endtime = end_of_study)

out_single_arm <- survival_adapt(
  hazard_treatment = ht_single,
  hazard_control = NULL,
  cutpoints = 0,
  N_total = 80,
  lambda = 8,
  lambda_time = 0,
  interim_look = 50,
  end_of_study = end_of_study,
  bin_prior = c(1, 1),
  bin_method = "quadrature",
  prop_loss = 0,
  alternative = "less",
  h0 = benchmark,
  Fn = 0.05,
  Sn = 0.90,
  prob_ha = 0.95,
  N_impute = 20,
  method = "bayes-bin",
  imputed_final = FALSE
)

out_single_arm
```

In this setting `est_final` is the posterior mean event probability in the
single arm, and `post_prob_ha` is the posterior probability that this event
probability is below the benchmark.

## Choosing `bin_method`

The posterior probability can be calculated in three ways:

- `bin_method = "mc"` draws from the beta posterior directly. Use `bin_N` to
  control the number of Monte Carlo draws.
- `bin_method = "normal"` uses a normal approximation to the posterior mean or
  treatment-control difference.
- `bin_method = "quadrature"` uses numerical integration for the two-arm
  posterior difference, and the closed-form beta CDF for single-arm designs.

The Monte Carlo option is usually the most flexible. The quadrature option is
deterministic and useful for small examples or checks. The normal approximation
is fastest, but should be used with care when sample sizes are small or event
probabilities are near 0 or 1.

## Operating characteristics

As with the survival methods, operating characteristics should be evaluated by
simulation. The `seed` argument makes the simulation reproducible, including
when `ncores` is greater than 1 on platforms that support parallel simulation.

```{r oc, eval=FALSE}
out_power <- sim_trials(
  N_trials = 1000,
  hazard_treatment = ht,
  hazard_control = hc,
  cutpoints = 0,
  N_total = 120,
  lambda = 10,
  lambda_time = 0,
  interim_look = 80,
  end_of_study = end_of_study,
  bin_prior = c(1, 1),
  bin_method = "quadrature",
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0,
  alternative = "less",
  h0 = 0,
  Fn = 0.05,
  Sn = 0.90,
  prob_ha = 0.95,
  N_impute = 20,
  method = "bayes-bin",
  imputed_final = FALSE,
  ncores = 2,
  seed = 5107
)

out_null <- update(out_power, hazard_treatment = hc, seed = 5108)

summarise_sims(list(out_power$sims, out_null$sims))
```

The binary endpoint model is still calibrated through the event-time simulator.
When the binary endpoint corresponds to event status by `end_of_study`, choose
hazards that reproduce clinically meaningful endpoint probabilities through
`prop_to_haz()` and `ppwe()`.
