---
title: "Calibrating stopping rules from complete trial paths"
author: "Graeme L. Hickey"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Calibrating stopping rules from complete trial paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(goldilocks)
```

Repeated threshold calibration can spend most of its time generating and analysing the same kinds of predictive completions. `sim_trial_paths()` performs these calculations once at every planned look. `apply_stopping_rules()` then evaluates different `Fn`, `Sn`, `Qn`, and `prob_ha` values using the saved evidence, without further imputation or model fitting.

A bank represents one fixed data-generating scenario and analysis specification. Separate banks are required for the null, alternatives, or changes to the priors, analysis method, endpoint horizon, allocation, maximum sample size, look schedule, or Monte Carlo settings. This workflow supplements `sim_trials()`, which remains useful for evaluating individual designs.

## What is saved

At each interim look, the bank retains the success score from every predictive completion of the currently enrolled cohort and the maximum cohort. A score is a posterior probability for a Bayesian analysis or $1-p$ for a frequentist analysis. For a candidate completed-data threshold $c$, the predictive probability is reconstructed as

$$\widehat P(c) = \frac{1}{M}\sum_{m=1}^{M} I(s_m > c),$$

where $M$ is `N_impute`. Retaining these scores is necessary because changing `prob_ha` changes the predictive probabilities themselves. A trace containing only the predictive probabilities for one threshold cannot support this recalibration.

The bank also retains the final analysis and calendar-time measures for each possible stopping cohort. If accrual stops at 60 participants for expected success, the final outcome comes from those 60 participants after follow-up. The final result from the maximum cohort is reserved for trials that continue to maximum N and for the separate counterfactual comparison.

Interim calculations use calendar-masked data. The eventual outcomes used for cohort final analyses are not made available to the interim predictions. Dropout and final imputation follow the same rules as `survival_adapt()`.

## Create null and alternative banks

The example has four analyses: interim enrollment looks at 40, 60, and 80 participants, followed by the final analysis of 100 participants after follow-up. Accrual is six participants per month, follow-up is 12 months, allocation is equal, and there is no dropout. The control event risk is 30% at 12 months; the alternative treatment risk is 20%. The primary analysis is a Bayesian beta-binomial comparison of the treatment-minus-control event risk, using a uniform Beta(1, 1) prior in each arm and numerical quadrature.

The small simulation counts below demonstrate the workflow. They do not provide enough precision to establish type I error control or select a design.

```{r simulation-banks}
fixed_design <- list(
  hazard_control = prop_to_haz(0.30, endtime = 12),
  N_total = 100,
  interim_look = c(40, 60, 80),
  lambda = 6,
  lambda_time = NULL,
  end_of_study = 12,
  cutpoints = NULL,
  generation_cutpoints = NULL,
  prior_surv = c(0.1, 0.1),
  prior_surv_final = c(0.1, 0.1),
  block = 2,
  rand_ratio = c(control = 1, treatment = 1),
  prop_loss = c(control = 0, treatment = 0),
  alternative = "less",
  h0 = 0,
  method = "bayes-bin",
  prior_bin = c(1, 1),
  bin_method = "quadrature",
  imputed_final = FALSE,
  N_impute = 10,
  N_mcmc = 1000,
  mc_conf_level = 0.95,
  ncores = 1,
  backend = "sequential"
)

null_args <- c(fixed_design, list(
  hazard_treatment = prop_to_haz(0.30, endtime = 12),
  N_trials = 40, seed = 8201
))
alternative_args <- c(fixed_design, list(
  hazard_treatment = prop_to_haz(0.20, endtime = 12),
  N_trials = 40, seed = 8202
))

banks <- list(
  null = do.call(sim_trial_paths, null_args),
  alternative = do.call(sim_trial_paths, alternative_args)
)
banks$null
```

`N_impute` is part of the implemented stopping algorithm. It is fixed within each bank and should remain fixed when validating a selected candidate. `N_mcmc` likewise remains fixed for Monte Carlo Bayesian analyses; it is unused for the quadrature calculation in this example. Increasing `N_trials` improves operating-characteristic precision without changing the decision algorithm.

## Evaluate candidate rules

This grid varies futility, expected-success, and completed-data thresholds. Immediate success remains disabled with `Qn = 1`. Scalars are applied at every look; vectors of length three would specify separate interim boundaries for this design.

```{r candidates}
grid <- expand.grid(
  Fn = c(0.05, 0.10),
  Sn = c(0.85, 0.95),
  Qn = 1,
  prob_ha = c(0.975, 0.99)
)
rules <- lapply(seq_len(nrow(grid)), function(i) as.list(grid[i, ]))
names(rules) <- paste0("rule_", seq_along(rules))

results <- lapply(banks, function(bank) {
  lapply(rules, function(rule) {
    do.call(apply_stopping_rules, c(list(paths = bank), rule))
  })
})
null_summary <- summarise_sims(results$null)
alternative_summary <- summarise_sims(results$alternative)
knitr::kable(cbind(candidate = names(rules), grid), digits = 3)
```

The first stopping decision is terminal. Immediate success takes precedence over expected success, and either success decision takes precedence over futility. Expected success stops accrual and requires the corresponding cohort final analysis. Binding futility is an official failure even if the maximum-N counterfactual would succeed. All boundary comparisons are strict; equality does not cross a boundary.

Under the null, the `power` column is the type I error estimate. Its Monte Carlo interval reflects uncertainty from the finite number of simulated trials.

```{r null-results}
display <- null_summary[c(
  "scenario", "n_used", "power", "power_mc_lower", "power_mc_upper", "mean_N"
)]
names(display)[1:5] <- c("Candidate", "Trials", "Type I error", "Lower", "Upper")
knitr::kable(display, digits = 3)
```

## Two distinct futility measures

`stop_futility` estimates the probability of binding futility under the specified data-generating scenario. Under the alternative, it answers how often that design stops for futility despite the assumed treatment effect.

`futility_and_success_at_max` estimates the joint probability of binding futility and success had the same simulated trial continued to maximum N without interim stopping. The counterfactual uses the same final analysis and the candidate's `prob_ha`. Its denominator is all available joint-outcome comparisons, not only trials stopping for futility. This is not the net power loss relative to a fixed maximum-N design: some adaptive trials can succeed when their maximum-N counterfactual would fail.

```{r alternative-results}
display <- alternative_summary[c(
  "scenario", "power", "stop_futility", "futility_and_success_at_max", "mean_N"
)]
names(display) <- c(
  "Candidate", "Power", "Futility", "Futility and max-N success", "Mean N"
)
knitr::kable(display, digits = 3)
```

## Audit decisions and failures

Request a candidate trace when inspecting individual decisions. It includes only reached looks. The original bank retains all hypothetical looks.

```{r inspect-path}
selected_example <- do.call(apply_stopping_rules, c(
  list(paths = banks$alternative, return_trace = TRUE), rules[[1]]
))
selected_example$traces[selected_example$traces$trial == 1,
  c("trial", "planned_N", "ppp_stop_now", "ppp_success_at_max", "decision")]
banks$alternative$trials[[1]]$finals[
  c("planned_N", "post_prob_ha", "analysis_ready_time")
]
```

Inspect `bank$failures` for failures by trial, look, and calculation. One failed predictive replicate invalidates its entire cohort predictive probability; the implementation does not change the denominator by dropping that replicate. Later calculations are still attempted. A candidate only loses a trial when a calculation required by its decision path is unavailable. Its own `result$failures` records those exclusions.

A failed maximum-N final analysis does not invalidate an otherwise known early adaptive outcome. For a futility stop it makes `futility_and_success_at_max` unavailable. The summary reports `futility_and_success_at_max_n_used` and `futility_and_success_at_max_n_missing`; its interval uses available comparisons only. Missing comparisons may be informative and should be reviewed alongside the estimated rates.

## Compare candidates and validate independently

All candidates within a scenario share the same simulated trials. Marginal Monte Carlo intervals are suitable for individual operating characteristics, but differences between candidates require paired calculations. Align results using the original `trial` identifiers, particularly if candidates have different failures.

```{r paired-comparison}
a <- results$alternative[[1]]$sims
b <- results$alternative[[2]]$sims
paired <- merge(a[c("trial", "trial_success")], b[c("trial", "trial_success")],
  by = "trial", suffixes = c("_a", "_b"))
difference <- as.numeric(paired$trial_success_b) - as.numeric(paired$trial_success_a)
c(difference = mean(difference), mcse = sd(difference) / sqrt(length(difference)))
```

Prespecify calibration targets and candidate-selection criteria. For example, require the upper null Monte Carlo bound to be below the type I error target, then assess power, futility, enrollment, and duration among qualifying candidates. A composite null may require several null banks. Screening many candidates on one bank does not provide independent confirmation of the selected rule.

The following template generates an independent validation bank after a candidate has been selected using an adequately sized screening study. `selected_rule` must contain that candidate's four threshold settings. The Monte Carlo settings governing interim decisions remain unchanged.

```{r independent-validation, eval = FALSE}
validation_args <- null_args
validation_args$N_trials <- 10000
validation_args$seed <- 9201
validation_bank <- do.call(sim_trial_paths, validation_args)
validation_result <- do.call(apply_stopping_rules, c(
  list(paths = validation_bank), selected_rule
))
summarise_sims(validation_result)
```

## Save and process batches

Save the complete bank with `saveRDS()`; exporting only scores to CSV loses metadata and posterior Monte Carlo counts. For large studies, generate batches with one explicit seed and consecutive `trial_offset` values. Each batch is held in memory and can be saved separately. Keeping all other settings fixed produces the same paths and trial identifiers as a single large call.

```{r save-batches, eval = FALSE}
for (batch in 0:9) {
  batch_args <- null_args
  batch_args$N_trials <- 1000
  batch_args$trial_offset <- batch * 1000L
  bank <- do.call(sim_trial_paths, batch_args)
  saveRDS(bank, sprintf("null-paths-%02d.rds", batch + 1L))
}
```

Evaluate each saved batch separately when memory is limited. Combine trial outcomes and failure records before calculating overall operating characteristics; do not average batch confidence limits. Retain the requested trial count, scenario, seed, candidate rules, and original trial identifiers when combining results.

The same explicit seed produces identical bank paths across sequential, fork, and PSOCK backends where supported, and preserves the caller's random-number state. The bank uses separate streams for each trial and calculation stage. Its seeded results therefore need not match `sim_trials()`, whose existing draw scheduling is preserved. Rule application itself consumes no random numbers.
