---
title: "Inspecting adaptive decision paths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspecting adaptive decision paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(5701)
```

```{r setup, message = FALSE}
library(goldilocks)
```

An adaptive trial is easier to assess when the final result can be connected back to the interim decisions that led to it. By default, `survival_adapt()` returns its compact, one-row final summary. Set `return_trace = TRUE` to request a `goldilocks_trial` object with that summary and a tidy row for each interim look.

## A traced trial

This small Bayesian survival design has two interim looks. The treatment arm is assumed to have a lower cumulative failure probability by 24 months.

```{r traced-trial}
end_of_study <- 24
hazard_control <- prop_to_haz(c(0.20, 0.35), c(0, 12), end_of_study)
hazard_treatment <- prop_to_haz(c(0.12, 0.24), c(0, 12), end_of_study)

trial <- survival_adapt(
  hazard_treatment = hazard_treatment,
  hazard_control = hazard_control,
  cutpoints = c(0, 12),
  N_total = 80,
  lambda = 8,
  lambda_time = 0,
  interim_look = c(40, 60),
  end_of_study = end_of_study,
  prior = c(0.1, 0.1),
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0.05,
  alternative = "less",
  h0 = 0,
  Fn = c(0.05, 0.05),
  Sn = c(0.95, 0.90),
  prob_ha = 0.95,
  N_impute = 20,
  N_mcmc = 20,
  method = "bayes-surv",
  empty_interval = "prior",
  return_trace = TRUE
)

trial
```

The summary element is the familiar final analysis output. The trace element is an audit trail of the interim path.

```{r trace-table}
trial$summary
trial$trace
summarise_trial_trace(trial)
```

For each completed look, `ppp_stop_now` is the predictive probability of success if enrollment stops at that look. It is compared with `success_threshold`. `ppp_success_at_max` is the predictive probability of success if enrollment continues to the maximum sample size. It is compared with `futility_threshold`. The decision column records whether the design continued, stopped for expected success, or stopped for futility.

The trace does not retain posterior draws or completed imputed data sets. That keeps a traced trial small enough to inspect and share while leaving the underlying random-number path unchanged.

## Visualizing the path

```{r trace-plot, fig.width = 7, fig.height = 8}
plot_trial_trace(trial)
```

The first two panels show the two predictive probabilities alongside their decision thresholds. The final panel shows enrollment and observed events by arm at each look. Warnings raised during a look, such as empty-interval handling, are recorded in `warning_messages` and remain visible as ordinary R warnings.

## Summarizing many simulated trials

Traces are intended for examining individual trial paths. `sim_trials()` keeps its compact result data frame, which is more suitable for large operating characteristic simulations. `plot_sim_stopping()` provides a quick visual summary of stopping outcomes and enrolled sample sizes.

```{r simulation-summary, eval = FALSE}
sims <- sim_trials(
  hazard_treatment = hazard_treatment,
  hazard_control = hazard_control,
  cutpoints = c(0, 12),
  N_total = 80,
  lambda = 8,
  lambda_time = 0,
  interim_look = c(40, 60),
  end_of_study = end_of_study,
  prior = c(0.1, 0.1),
  block = 2,
  rand_ratio = c(1, 1),
  prop_loss = 0.05,
  alternative = "less",
  h0 = 0,
  Fn = c(0.05, 0.05),
  Sn = c(0.95, 0.90),
  prob_ha = 0.95,
  N_impute = 20,
  N_mcmc = 20,
  N_trials = 500,
  method = "bayes-surv",
  seed = 5702
)

summarise_sims(sims$sims)
plot_sim_stopping(sims)
```

For reproducible simulations, set seed in `sim_trials()`. The per-trial random streams make the compact simulation results reproducible across supported parallel backends. For a detailed explanation of the decision algorithm and calibration, see the "Technical details of the Goldilocks design" vignette.