---
title: "Applying predictive decisions to an observed interim data cut"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applying predictive decisions to an observed interim data cut}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(goldilocks)
```

`survival_adapt()` simulates a complete adaptive trial. `evaluate_interim()`
instead applies the same predictive decision calculation to one observed
interim data cut. It does not generate event, censoring, or enrollment times,
and it does not retain a participant-level copy of the input in its result.

## Preparing the data cut

The input has one row for every enrolled participant and uses a common time
unit throughout. Calendar-time `enrollment` is measured from first participant
randomization at time zero. Subject-level `time` is measured from that
participant's randomization.

The status values make the reason for incomplete follow-up explicit:

| `status` | Interpretation |
|---|---|
| `event` | The endpoint event has been observed. |
| `complete` | Event-free follow-up reached `end_of_study`. |
| `pending` | The participant remains under follow-up. |
| `censored` | Follow-up ended early without an event. |

A blinded data-management team can prepare enrollment, follow-up, event, and
status fields without access to treatment assignments:

```{r blinded-data}
blinded_cut <- data.frame(
  id = 1:8,
  enrollment = 0:7,
  time = c(8, 7, 6, 5, 4, 3, 2, 1),
  event = c(1, 0, 0, 1, 0, 0, 0, 0),
  status = c(
    "event", "pending", "censored", "event",
    "pending", "pending", "pending", "pending"
  )
)

blinded_cut
```

The calculation itself is arm-specific. An independent unblinded statistician
or service therefore joins the treatment assignment before running it. The
participant-level input need not be distributed with the aggregate result.

```{r unblinded-data}
randomization_export <- data.frame(
  id = 1:8,
  treatment = c(0, 1, 0, 1, 0, 1, 0, 1)
)

interim_cut <- blinded_cut
interim_cut$treatment <- randomization_export$treatment[
  match(interim_cut$id, randomization_export$id)
]
interim_cut <- interim_cut[
  c("id", "treatment", "enrollment", "time", "event", "status")
]
```

## Evaluating the look

Suppose this is the second planned look of a trial with maximum sample size 12,
equal randomization, and ten time units of follow-up per participant:

```{r evaluate-look}
interim_result <- evaluate_interim(
  data = interim_cut,
  data_cut = 8,
  look = 2,
  N_total = 12,
  end_of_study = 10,
  rand_ratio = c(control = 1, treatment = 1),
  alternative = "less",
  Fn = 0.05,
  Sn = 0.90,
  prob_ha = 0.95,
  N_impute = 20,
  seed = 20260831
)

interim_result
interim_result$decision
interim_result$monte_carlo
```

At the maximum sample size, equal randomization implies six participants per
arm. Four have already accrued in each arm, so the function internally adds two
potential future participants per arm for the maximum-sample calculation:

```{r allocation-diagnostics}
interim_result$diagnostics$target_allocation
interim_result$diagnostics$current_allocation
interim_result$diagnostics$potential_accruals
```

No block size or future randomization sequence is needed. The design must have
an exactly allocable maximum sample size: for example, a 1:2 ratio requires
`N_total` to be divisible by three. Observed enrollment in either arm cannot
already exceed its implied maximum.

## Reviewing and retaining the audit trail

The decision uses the Monte Carlo point estimates. The standard errors and
exact one-sided bounds describe finite-draw uncertainty but do not change the
decision. The one-row trace uses the same schema as a simulated adaptive-trial
trace and can be summarized or plotted with the existing helpers:

```{r audit-trace}
summarise_trial_trace(interim_result)
interim_result$trace
```

The metadata records the evaluated design, package version, data cut, fixed
time-origin convention, and random-number policy. Supplying `seed` makes the
calculation reproducible while preserving the caller's random-number state.
With `seed = NULL`, the calculation instead uses and advances the current R
random-number state.
