---
title: "ThermoCool AF: immediate success from predictive probability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ThermoCool AF: immediate success from predictive probability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
set.seed(1030236)
```

```{r setup, message = FALSE}
library(goldilocks)
```

The ThermoCool AF trial is a useful published example of a Bayesian design
that used the same predictive probability to distinguish between stopping
accrual and declaring immediate trial success. The trial compared radiofrequency
catheter ablation with antiarrhythmic drug therapy (ADT) in patients with
symptomatic paroxysmal atrial fibrillation. It was registered as
[NCT00116428](https://clinicaltrials.gov/study/NCT00116428), reported by
[Wilber et al. (2010)](https://jamanetwork.com/journals/jama/fullarticle/185277),
and supported FDA premarket approval supplement
[P030031/S011](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfpma/pma.cfm?id=P030031S011).

This vignette uses the trial to explain `Qn`, the upper boundary for immediate
success in `goldilocks`. It is a **ThermoCool-inspired package example, not an
exact reconstruction or regulatory validation**. The public sources disclose
the principal decision boundaries and longitudinal imputation model, but some
simulation inputs are unavailable. In addition, the package rule deliberately
differs from the trial protocol in its futility statistic, comparison operators,
and treatment of a separate information gate for an early claim.

## Trial and endpoint

ThermoCool AF was a prospective, multicenter, randomized, unblinded trial.
Patients were allocated within site in blocks of 11: seven to catheter ablation
and four to ADT. The primary effectiveness endpoint was freedom from
protocol-defined treatment failure during comparable nine-month evaluation
periods. Treatment failure included documented symptomatic paroxysmal atrial
fibrillation; the protocol also classified specified procedural, medication,
and intolerance outcomes as failures.

The planned maximum sample size was 230. Decision-capable interim analyses were
planned after 150, 175, and 200 accrued patients. Trial success was defined by

$$
\Pr(p_T > p_C \mid \mathcal D) \ge 0.98,
$$

where $p_T$ and $p_C$ are the nine-month chronic-success probabilities in the
ablation and ADT groups. The final posterior criterion was selected through
simulation to control the one-sided type I error at no more than 0.025.

The trial began with a fixed-sample frequentist design and was amended while
accrual was underway. FDA's advisory briefing reports that the 106 patients
already accrued were treated as a non-stopping look in the operating-characteristic
simulations, with a statistical penalty. This historical feature is important
to the reported calibration, but it is not an actionable `interim_look` in the
package example below.

## The reported stopping rules

At a planned interim analysis, the trial calculated the predictive probability
that the final posterior criterion would be met after currently enrolled
patients completed follow-up. In package notation this is $P_{n,\ell}$. The
reported rules were:

-   If $P_{n,\ell} \ge 0.99$, stop and make an early claim of success.
-   Otherwise, stop accrual for expected success if $P_{n,\ell} \ge 0.90$ at
    $n=150$, or $P_{n,\ell} \ge 0.80$ at $n=175$ or $200$, and continue
    follow-up.
-   Stop for futility if **both** $P_{n,\ell}<0.01$ and
    $P_{n_{\max},\ell}<0.01$.
-   Otherwise, continue accrual.

The JAMA methods summarize the 0.99 rule as an immediate claim at an interim
analysis. The FDA advisory materials provide an operational detail: after
accrual stopped, the early-claim analysis was performed once either 4.5 months
had elapsed or at least 50% of enrolled patients had complete effectiveness
outcomes, whichever came first. The public record reports that this information
condition was satisfied when the first analysis was performed.

In practice, the first planned 150-patient analysis occurred after 160 patients
had been enrolled because of operational timing; 148 were in the effectiveness
analysis. The predictive probability was greater than 0.999, 50% of enrollees
had an effectiveness determination, and the trial declared early success.
Seven more patients were enrolled before shutdown was completed, producing 167
randomized patients in the updated report.

## The four-way `goldilocks` rule

The package enhancement keeps predictive probability as the key interim
statistic but uses the following ordered rule:

$$
d_\ell =
\begin{cases}
\text{immediate success}, & P_{n,\ell} > Q_\ell, \\
\text{stop accrual for expected success},
  & S_\ell < P_{n,\ell} \le Q_\ell, \\
\text{binding futility},
  & P_{n,\ell} \le S_\ell
    \text{ and } P_{n_{\max},\ell} < F_\ell, \\
\text{continue}, & \text{otherwise.}
\end{cases}
$$

```{r decision-flow, echo = FALSE, message = FALSE, out.width = "100%"}
DiagrammeR::grViz("
digraph thermocool_rule {
  graph [rankdir = TB, bgcolor = transparent, ranksep = 0.50, nodesep = 0.35]
  node [shape = box, style = rounded, fontname = Helvetica, fontsize = 13,
        margin = 0.11]
  edge [fontname = Helvetica, fontsize = 11]

  start [label = 'At look l, calculate\nP[n,l] and P[nmax,l]']
  q [shape = diamond, label = 'P[n,l] > Q[l]?']
  immediate [label = 'Stop\nDeclare immediate success']
  s [shape = diamond, label = 'P[n,l] > S[l]?']
  expected [label = 'Stop accrual\nComplete follow-up, then analyze']
  f [shape = diamond, label = 'P[nmax,l] < F[l]?']
  futile [label = 'Stop for futility\nBinding decision']
  continue [label = 'Continue accrual\nTo the next look']

  start -> q
  q -> immediate [label = 'Yes']
  q -> s [label = 'No: P[n,l] <= Q[l]']
  s -> expected [label = 'Yes']
  s -> f [label = 'No']
  f -> futile [label = 'Yes']
  f -> continue [label = 'No']
}
", width = "100%", height = "670px")
```

Decision order matters. An immediate-success crossing is terminal even if a
futility boundary also appears crossed. Equality with `Qn` does not declare
immediate success, and equality with `Sn` does not stop accrual for expected
success. `Qn = 1` disables immediate-success stopping at a look because a
predictive probability cannot be strictly greater than one. The package
requires `Sn <= Qn` at every look; setting them equal removes the
expected-success interval at that look.

For the ThermoCool-inspired mapping, the threshold vectors are:

```{r stopping-thresholds}
N_total <- 230
interim_look <- c(150, 175, 200)

Qn <- rep(0.99, length(interim_look))
Sn <- c(0.90, 0.80, 0.80)
Fn <- rep(0.01, length(interim_look))
```

These values reproduce the reported numerical boundaries, but not every
protocol detail. The package uses strict upper-boundary comparisons, whereas
the publications describe "at least" 0.99, 0.90, or 0.80. More importantly,
the package's requested futility rule uses only $P_{n_{\max},\ell}<F_\ell$;
ThermoCool AF required both predictive probabilities below 0.01. The package
also evaluates `Qn` at each scheduled look without a separate 4.5-month/50%
information gate.

## Source and status of modeled inputs

The labels in the following table are used consistently throughout the
vignette:

-   **reported**: stated numerically in a public primary source;
-   **inferred**: calculated from, or used to encode, reported information;
-   **assumed**: selected for this runnable package example;
-   **unavailable**: needed for exact reconstruction but not publicly supplied
    or not representable in the current package interface.

```{r input-audit, echo = FALSE}
input_audit <- data.frame(
  Input = c(
    "`N_total`",
    "`interim_look`",
    "`Qn`",
    "`Sn`",
    "`Fn`",
    "Futility statistic",
    "Early-claim information gate",
    "Non-stopping 106-patient look",
    "`prob_ha`",
    "`method`, `alternative`, `h0`",
    "`prior_bin`",
    "`cutpoints`, `end_of_study`",
    "Sponsor predictive hazard prior",
    "`prior_surv`, `prior_surv_final`",
    "`block`, `rand_ratio`",
    "`lambda`, `lambda_time`",
    "Data-generating hazards",
    "Sponsor planning scenarios",
    "`prop_loss`",
    "`imputed_final`",
    "`N_impute`",
    "Evaluated `N_trials`"
  ),
  Value = c(
    "230",
    "150, 175, 200",
    "0.99 at each look",
    "0.90, 0.80, 0.80",
    "0.01 at each look",
    "Package uses P[nmax,l] only",
    "Not represented",
    "Not represented as an actionable look",
    "0.98",
    "bayes-bin, less, 0",
    "Beta(1, 1) in both arms",
    "0.5 and 2 months; 9-month horizon",
    "Three arm-specific piecewise rates with a hierarchical prior",
    "Fixed Gamma(1, 1) approximation in each interval and arm",
    "11; control 4 : treatment 7",
    "0.83, 3.50, 8.92 patients/month; changes at months 12 and 24",
    "June 2008 failures divided by exposure within each interval",
    "Not publicly available in sufficient detail",
    "0.05 in each arm",
    "TRUE",
    "100 for one trial; 40 for small OC simulations",
    "50 per scenario"
  ),
  Status = c(
    "reported", "reported", "reported", "reported", "reported",
    "assumed", "reported", "reported", "reported", "inferred",
    "reported", "reported", "reported", "assumed", "reported",
    "inferred", "inferred", "unavailable", "assumed", "inferred",
    "assumed", "assumed"
  ),
  `Source and mapping` = c(
    "JAMA Statistical Methods and FDA advisory briefing",
    "JAMA Statistical Methods and FDA advisory briefing",
    "Reported early-claim boundary; package applies its strict > rule",
    "Reported expected-success accrual boundaries",
    "Reported numeric futility boundary",
    "The requested enhancement differs from ThermoCool's P[n,l] and P[nmax,l] rule",
    "FDA briefing and transcript report 4.5 months or 50% endpoint-complete",
    "FDA requested it after the mid-trial amendment; zero probability of stopping",
    "Historical success was at least 0.98; the package classifies success using its strict > comparison",
    "Failure is the modeled binary event, so benefit is a lower treatment failure probability",
    "FDA advisory transcript and sponsor briefing",
    "JAMA and FDA describe breaks at 2 weeks/0.5 months and 2 months",
    "FDA briefing discloses Exp(rate = 1) priors on the Gamma hyperparameters, but not a directly reproducible fixed package prior",
    "Assumed plug-in approximation that sets both Gamma hyperparameters to the mean, 1, of their disclosed Exp(rate = 1) hyperpriors; not the sponsor's hierarchical prior",
    "JAMA Study Design",
    "Derived from FDA's 11 patients after year 1, 53 after year 2, and 160 near year 3; not a protocol accrual generator",
    "Derived from the FDA SSED Table 8; descriptive observed-data scenario, not a planning assumption",
    "Public materials state that operating characteristics were simulated but redact or omit enough inputs to prevent exact reproduction",
    "Illustrative assumption; random censoring is not equivalent to the eight randomized patients who did not receive assigned treatment",
    "JAMA reports multiple imputation for incomplete outcomes; the available package analysis is an approximation",
    "Illustrative choices; the sponsor's predictive draw count is unavailable",
    "Illustrative choice; the sponsor briefing reports 10,000 simulated trials per scenario"
  ),
  check.names = FALSE
)

knitr::kable(input_audit, format = "pipe")
```

## Mapping the longitudinal model

The completed endpoint is binary, but its status is learned over time. The
trial predicted unknown nine-month outcomes with an arm-specific
piecewise-exponential time-to-failure model. The three intervals were
0--0.5 months, 0.5--2 months, and 2--9 months. After imputation, a patient's
failure time was reduced to a binary chronic-failure indicator, and the
completed-data analysis compared arm-specific success probabilities.

`goldilocks` represents this structure with `method = "bayes-bin"` and
`binary_imputation = "event-time"`. The package is parameterized on event
probability, so the code models chronic **failure** rather than chronic success:

$$
p_{\text{failure}} = 1 - p_{\text{chronic success}}.
$$

Consequently, superiority of ablation is encoded as

$$
\Pr(p_{T,\text{failure}} - p_{C,\text{failure}} < 0 \mid \mathcal D) > 0.98,
$$

using `alternative = "less"` and `h0 = 0`. Independent uniform priors on the
published chronic-success probabilities become the same independent
`Beta(1, 1)` priors after complementing to failure probabilities.

```{r analysis-model}
analysis_cutpoints_month <- c(0.5, 2)
effectiveness_horizon_month <- 9

prior_bin <- c(1, 1)

# Assumed plug-in approximation using the means of the disclosed Exp(1)
# hyperpriors; this is not the sponsor's hierarchical hazard prior.
prior_surv <- c(shape = 1, rate = 1)
```

The FDA summary reports observed exposure and failure counts for the same three
intervals in its June 2008 analysis. Dividing failures by exposure produces a
fully documented descriptive generator:

```{r descriptive-hazards}
fda_interval_data <- data.frame(
  arm = rep(c("treatment", "control"), each = 3),
  interval = rep(c("0--0.5", "0.5--2", "2--9"), 2),
  exposure_months = c(40.21, 104.17, 413.09, 23.27, 54.21, 90.46),
  failures = c(26, 3, 7, 13, 14, 20)
)
fda_interval_data$hazard_per_month <- with(
  fda_interval_data,
  failures / exposure_months
)

knitr::kable(fda_interval_data, digits = 4)

hazard_treatment <- subset(
  fda_interval_data,
  arm == "treatment"
)$hazard_per_month
hazard_control <- subset(
  fda_interval_data,
  arm == "control"
)$hazard_per_month
```

These are post-trial empirical rates, not prospective design assumptions.
They also approximate failures assigned at the start of the evaluation window
with a continuous hazard. They are used only to create a strong-benefit scenario
that resembles the information pattern in the public record.

The event-free probabilities implied directly by these rates are:

```{r implied-probabilities}
interval_length_month <- diff(c(
  0,
  analysis_cutpoints_month,
  effectiveness_horizon_month
))

implied_chronic_success <- data.frame(
  Arm = c("Catheter ablation", "ADT control"),
  `Implied nine-month chronic-success probability` = c(
    exp(-sum(hazard_treatment * interval_length_month)),
    exp(-sum(hazard_control * interval_length_month))
  ),
  check.names = FALSE
)

knitr::kable(implied_chronic_success, digits = 3)
```

The JAMA paper reports Kaplan-Meier chronic-success estimates of 0.66 and 0.16.
The values above need not equal those estimates: they are obtained by inserting
raw interval rates into a piecewise-exponential distribution, without the
sponsor's full posterior calculation, protocol handling of time-zero failures,
or analysis-population rules.

## An observed-history accrual approximation

Accrual speed determines how much nine-month endpoint information is available
at an enrollment-triggered look. FDA reports 11 accrued patients after one
year, 53 after two years, and 160 near the third year. The following piecewise
rates preserve those increments approximately:

```{r accrual-model}
enrollment_rate_per_month <- c(
  year_1 = (11 - 1) / 12,
  year_2 = (53 - 11) / 12,
  year_3 = (160 - 53) / 12
)
enrollment_rate_change_month <- c(12, 24)

data.frame(
  `Trial-calendar interval` = c("0--12 months", "12--24 months", "24+ months"),
  `Patients per month` = enrollment_rate_per_month,
  check.names = FALSE
)
```

This is an inferred operational approximation, not a reported Poisson accrual
model. `goldilocks` fixes the first enrollment at trial-calendar time zero, so
the first rate generates 10 additional expected arrivals by month 12, for 11
patients in total. The later rates use the reported cumulative increments. The
schedule intentionally captures the slow start and later acceleration because
a single average rate would give a different amount of follow-up at the first
look.

## One simulated trial

The following analysis simulates one trial replicate under the descriptive
FDA-rate scenario. It uses only 100 predictive imputations for illustration.
That is too few for regulatory calibration of boundaries as extreme as 0.99
and 0.01.

```{r one-trial}
set.seed(1030236)

thermocool_trial <- survival_adapt(
  hazard_treatment = hazard_treatment,
  hazard_control = hazard_control,
  cutpoints = analysis_cutpoints_month,
  N_total = N_total,
  lambda = enrollment_rate_per_month,
  lambda_time = enrollment_rate_change_month,
  interim_look = interim_look,
  end_of_study = effectiveness_horizon_month,
  prior_surv = prior_surv,
  prior_surv_final = prior_surv,
  prior_bin = prior_bin,
  bin_method = "quadrature",
  binary_imputation = "event-time",
  block = 11,
  rand_ratio = c(control = 4, treatment = 7),
  prop_loss = 0.05,
  alternative = "less",
  h0 = 0,
  Fn = Fn,
  Sn = Sn,
  Qn = Qn,
  prob_ha = 0.98,
  N_impute = 100,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE,
  return_trace = TRUE
)

knitr::kable(
  thermocool_trial$summary[, c(
    "N_enrolled",
    "ppp_success",
    "stop_immediate_success",
    "stop_expected_success",
    "stop_futility",
    "trial_success",
    "stopping_reason",
    "decision_time"
  )],
  digits = 3
)
```

`stop_immediate_success` is an official terminal success decision. It does not
wait for a later completed-follow-up analysis, so `post_prob_ha` and the final
effect estimate are `NA`. Binding futility is likewise an official terminal
failure. For compatibility, the package still attempts its historical
completed-follow-up diagnostic after futility; if that diagnostic cannot be
computed, `post_prob_ha` and `est_final` are `NA` and `trial_success` remains
`FALSE`.

The trace makes the ordered rule auditable:

```{r one-trial-trace}
knitr::kable(
  thermocool_trial$trace[, c(
    "look",
    "planned_N",
    "ppp_stop_now",
    "immediate_success_threshold",
    "success_threshold",
    "ppp_success_at_max",
    "futility_threshold",
    "decision"
  )],
  digits = 3
)
```

Only looks actually reached appear in the trace. An immediate-success decision
prevents all subsequent looks.

## Small operating-characteristic demonstration

For simulation, collect common inputs in a named list and vary only the
data-generating hazards. The first scenario uses the FDA-derived descriptive
rates. The second assigns the control rates to both arms and is therefore a
simple no-treatment-effect scenario.

```{r small-oc}
thermocool_design <- list(
  cutpoints = analysis_cutpoints_month,
  N_total = N_total,
  lambda = enrollment_rate_per_month,
  lambda_time = enrollment_rate_change_month,
  interim_look = interim_look,
  end_of_study = effectiveness_horizon_month,
  prior_surv = prior_surv,
  prior_surv_final = prior_surv,
  prior_bin = prior_bin,
  bin_method = "quadrature",
  binary_imputation = "event-time",
  block = 11,
  rand_ratio = c(control = 4, treatment = 7),
  prop_loss = 0.05,
  alternative = "less",
  h0 = 0,
  Fn = Fn,
  Sn = Sn,
  Qn = Qn,
  prob_ha = 0.98,
  N_impute = 40,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE,
  N_trials = 50,
  ncores = 2,
  return_trace = TRUE
)

thermocool_descriptive <- do.call(sim_trials, c(
  thermocool_design,
  list(
    hazard_treatment = hazard_treatment,
    hazard_control = hazard_control,
    seed = 1030236
  )
))

thermocool_null <- do.call(sim_trials, c(
  thermocool_design,
  list(
    hazard_treatment = hazard_control,
    hazard_control = hazard_control,
    seed = 1030237
  )
))

oc_small <- summarise_sims(list(
  "FDA-rate illustration" = thermocool_descriptive,
  "Equal-arm null" = thermocool_null
))

oc_display <- oc_small[, c(
  "scenario",
  "n_analyzed",
  "power",
  "stop_immediate_success",
  "stop_success",
  "stop_futility",
  "stop_max_N",
  "mean_N"
)]
names(oc_display)[names(oc_display) == "stop_success"] <-
  "stop_expected_success"

knitr::kable(oc_display, digits = 3)
```

With only 50 trials and 40 imputations per look, these estimates have
substantial Monte Carlo error. They illustrate the mutually exclusive stopping
categories; they do not estimate the trial's reported operating
characteristics. `summarise_sims()` uses `stop_success` for stopping accrual for
expected success and reports immediate success separately as
`stop_immediate_success`.

The no-effect scenario provides a more informative view of the four decision
regions than the deliberately strong FDA-rate scenario:

```{r stopping-plot, fig.width = 7, fig.height = 5}
plot_sim_stopping(thermocool_null)
```

```{r decision-plot, fig.width = 9, fig.height = 7}
plot_sim_decisions(thermocool_null)
```

In a decision plot, `Qn` adds a second horizontal boundary above `Sn`.
Observations above `Qn` are immediate successes. Observations between `Sn` and
`Qn` stop accrual for expected success. At or below `Sn`, `Fn` separates binding
futility from continued accrual.

## Operating-characteristic evaluation

A design that permits immediate declaration of success must be calibrated as a
whole. Neither `Qn = 0.99` nor `prob_ha = 0.98` automatically controls type I
error in a new design. Calibration must reflect all looks, the final analysis,
the binding futility rule, accrual, incomplete follow-up, the predictive model,
and Monte Carlo error.

The following unevaluated template increases both the number of simulated
trials and the predictive-imputation count. The latter value is an analyst
choice, not a reported ThermoCool setting.

```{r full-validation, eval = FALSE}
thermocool_full_design <- modifyList(thermocool_design, list(
  N_trials = 10000,
  N_impute = 5000,
  ncores = 8,
  return_trace = FALSE
))

q_grid <- c(0.975, 0.99, 0.995, 1.00)

full_null <- lapply(seq_along(q_grid), function(i) {
  do.call(sim_trials, c(
    modifyList(thermocool_full_design, list(Qn = q_grid[i])),
    list(
      hazard_treatment = hazard_control,
      hazard_control = hazard_control,
      seed = 1031000 + i
    )
  ))
})
names(full_null) <- paste0("null: Qn = ", q_grid)

full_benefit <- lapply(seq_along(q_grid), function(i) {
  do.call(sim_trials, c(
    modifyList(thermocool_full_design, list(Qn = q_grid[i])),
    list(
      hazard_treatment = hazard_treatment,
      hazard_control = hazard_control,
      seed = 1032000 + i
    )
  ))
})
names(full_benefit) <- paste0("benefit: Qn = ", q_grid)

full_oc <- summarise_sims(c(full_null, full_benefit))
full_oc[, c(
  "scenario",
  "n_analyzed",
  "power",
  "power_mcse",
  "stop_immediate_success",
  "stop_immediate_success_mcse",
  "stop_success",
  "stop_futility",
  "stop_max_N",
  "mean_N",
  "mean_N_mcse"
)]
```

Useful additional scenarios include weaker treatment effects, different
control event rates, slower and faster accrual, arm-specific missingness,
alternative piecewise hazards, and violations of the predictive model. The
null design should be evaluated over a suitable nuisance-parameter range rather
than at a single equal-arm profile. Because a 0.99 boundary is estimated from
imputations, the predictive draw count should also be chosen so that decisions
near `Qn`, `Sn`, and `Fn` are sufficiently stable.

## What this example does and does not reproduce

The example preserves the central statistical idea:

-   one current-sample predictive probability drives two distinct upper
    decisions;
-   `Qn` declares success immediately, whereas `Sn` only stops accrual;
-   `Fn` acts on predictive success at the maximum sample size; and
-   all thresholds are evaluated in a prespecified order.

It does not reproduce the sponsor's complete analysis, hierarchical hazard
prior, simulation scenarios, treatment-specific evaluation-window origins,
site-stratified randomization sequences, protocol deviations, crossover,
analysis-population exclusions, operational overrun, or the original
two-predictive-probability futility requirement. It also does not recreate the
mid-trial design amendment or its statistical penalty. Those distinctions are
why the simulated numerical results should not be compared directly with the
trial's regulatory analysis.

## Primary sources

-   Wilber DJ, Pappone C, Neuzil P, et al. [Comparison of antiarrhythmic drug
    therapy and radiofrequency catheter ablation in patients with paroxysmal
    atrial fibrillation: a randomized controlled trial](https://jamanetwork.com/journals/jama/fullarticle/185277).
    *JAMA*. 2010;303(4):333-340. doi:10.1001/jama.2009.2029.
-   U.S. Food and Drug Administration. [Summary of Safety and Effectiveness
    Data, PMA P030031/S011](https://www.accessdata.fda.gov/cdrh_docs/pdf3/P030031S011B.pdf).
    The summary reports the approval analysis, longitudinal-model intervals,
    interval exposure and failures, and updated Bayesian results.
-   U.S. Food and Drug Administration. [Circulatory System Devices Panel,
    November 20, 2008 meeting materials](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfAdvisory/details.cfm?mtg=705).
    The archive links the FDA and sponsor briefing information, slides, panel
    questions, and transcript used for the design-history and early-claim
    details above.
-   U.S. Food and Drug Administration. [PMA supplement P030031/S011 approval
    record](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfpma/pma.cfm?id=P030031S011).
    FDA issued the decision on February 6, 2009 for the expanded paroxysmal
    atrial-fibrillation indication.
