---
title: "ThermoCool AF: immediate success from predictive probability"
output:
  rmarkdown::html_vignette:
    self_contained: false
    lib_dir: libs
    css: libs/vignette.css
vignette: >
  %\VignetteIndexEntry{ThermoCool AF: immediate success from predictive probability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("shared-vignette-resources.R")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
set.seed(1030236)
```

```{r setup, message = FALSE}
library(goldilocks)
```

The ThermoCool AF trial is a useful published example of a Bayesian design
that used the same predictive probability to distinguish between stopping
accrual and declaring immediate trial success. The trial compared radiofrequency
catheter ablation with antiarrhythmic drug therapy (ADT) in patients with
symptomatic paroxysmal atrial fibrillation. It was registered as
[NCT00116428](https://clinicaltrials.gov/study/NCT00116428), reported by
[Wilber et al. (2010)](https://pubmed.ncbi.nlm.nih.gov/20103757/),
and supported FDA premarket approval supplement
[P030031/S011](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfpma/pma.cfm?id=P030031S011).

This vignette uses the trial to explain `Qn`, the upper boundary for immediate
success in `goldilocks`. It is a **ThermoCool-inspired package example, not an
exact reconstruction or regulatory validation**. JAMA and the public FDA
record disclose the principal decision boundaries and longitudinal imputation
model, but omit or redact the numerical operating-characteristic scenarios.
The trial designers subsequently published the default failure-time generator,
accrual schedule, and scenario grid in
[Berry et al. (2010, Section 5.8, pp. 241--246)](https://doi.org/10.1201/EBK1439825488).
This example uses those published defaults. The complete regulatory sensitivity
suite remains unavailable, and the package rule deliberately differs from the
trial protocol in its futility statistic, comparison operators, and treatment
of a separate information gate for an early claim.

## Trial and endpoint

ThermoCool AF was a prospective, multicenter, randomized, unblinded trial.
Patients were allocated within site in blocks of 11: seven to catheter ablation
and four to ADT. The primary effectiveness endpoint was freedom from
protocol-defined treatment failure during comparable nine-month evaluation
periods. Treatment failure included documented symptomatic paroxysmal atrial
fibrillation; the protocol also classified specified procedural, medication,
and intolerance outcomes as failures.

The planned maximum sample size was 230. Decision-capable interim analyses were
planned after 150, 175, and 200 accrued patients. Trial success was defined by

$$
\Pr(p_T > p_C \mid \mathcal D) \ge 0.98,
$$

where $p_T$ and $p_C$ are the nine-month chronic-success probabilities in the
ablation and ADT groups. The final posterior criterion was selected through
simulation to control the one-sided type I error at no more than 0.025.

The trial began with a fixed-sample frequentist design and was amended while
accrual was underway. FDA's advisory briefing reports that the 106 patients
already accrued were treated as a non-stopping look in the operating-characteristic
simulations, with a statistical penalty. This historical feature is important
to the reported calibration, but it is not an actionable `interim_look` in the
package example below.

## The reported stopping rules

At a planned interim analysis, the trial calculated the predictive probability
that the final posterior criterion would be met after currently enrolled
patients completed follow-up. In package notation this is $P_{n,\ell}$. The
reported rules were:

-   If $P_{n,\ell} \ge 0.99$, stop and make an early claim of success.
-   Otherwise, stop accrual for expected success if $P_{n,\ell} \ge 0.90$ at
    $n=150$, or $P_{n,\ell} \ge 0.80$ at $n=175$ or $200$, and continue
    follow-up.
-   Stop for futility if **both** $P_{n,\ell}<0.01$ and
    $P_{n_{\max},\ell}<0.01$.
-   Otherwise, continue accrual.

The JAMA methods summarize the 0.99 rule as an immediate claim at an interim
analysis. The FDA advisory materials provide an operational detail: after
accrual stopped, the early-claim analysis was performed once either 4.5 months
had elapsed or at least 50% of enrolled patients had complete effectiveness
outcomes, whichever came first. The public record reports that this information
condition was satisfied when the first analysis was performed.

In practice, the first planned 150-patient analysis occurred after 160 patients
had been enrolled because of operational timing; 148 were in the effectiveness
analysis. The predictive probability was greater than 0.999, 50% of enrollees
had an effectiveness determination, and the trial declared early success.
Seven more patients were enrolled before shutdown was completed, producing 167
randomized patients in the updated report.

## The four-way `goldilocks` rule

The package enhancement keeps predictive probability as the key interim
statistic but uses the following ordered rule:

$$
d_\ell =
\begin{cases}
\text{immediate success}, & P_{n,\ell} > Q_\ell, \\
\text{stop accrual for expected success},
  & S_\ell < P_{n,\ell} \le Q_\ell, \\
\text{binding futility},
  & P_{n,\ell} \le S_\ell
    \text{ and } P_{n_{\max},\ell} < F_\ell, \\
\text{continue}, & \text{otherwise.}
\end{cases}
$$

```{r decision-flow, echo = FALSE, message = FALSE, out.width = "100%"}
DiagrammeR::grViz("
digraph thermocool_rule {
  graph [rankdir = TB, bgcolor = transparent, ranksep = 0.50, nodesep = 0.35]
  node [shape = box, style = rounded, fontname = Helvetica, fontsize = 13,
        margin = 0.11]
  edge [fontname = Helvetica, fontsize = 11]

  start [label = 'At look l, calculate\nP[n,l] and P[nmax,l]']
  q [shape = diamond, label = 'P[n,l] > Q[l]?']
  immediate [label = 'Stop\nDeclare immediate success']
  s [shape = diamond, label = 'P[n,l] > S[l]?']
  expected [label = 'Stop accrual\nComplete follow-up, then analyze']
  f [shape = diamond, label = 'P[nmax,l] < F[l]?']
  futile [label = 'Stop for futility\nBinding decision']
  continue [label = 'Continue accrual\nTo the next look']

  start -> q
  q -> immediate [label = 'Yes']
  q -> s [label = 'No: P[n,l] <= Q[l]']
  s -> expected [label = 'Yes']
  s -> f [label = 'No']
  f -> futile [label = 'Yes']
  f -> continue [label = 'No']
}
", width = "100%", height = "670px")
```

Decision order matters. An immediate-success crossing is terminal even if a
futility boundary also appears crossed. Equality with `Qn` does not declare
immediate success, and equality with `Sn` does not stop accrual for expected
success. `Qn = 1` disables immediate-success stopping at a look because a
predictive probability cannot be strictly greater than one. The package
requires `Sn <= Qn` at every look; setting them equal removes the
expected-success interval at that look.

For the ThermoCool-inspired mapping, the threshold vectors are:

```{r stopping-thresholds}
N_total <- 230
interim_look <- c(150, 175, 200)

Qn <- rep(0.99, length(interim_look))
Sn <- c(0.90, 0.80, 0.80)
Fn <- rep(0.01, length(interim_look))
```

These values reproduce the reported numerical boundaries, but not every
protocol detail. The package uses strict upper-boundary comparisons, whereas
the publications describe "at least" 0.99, 0.90, or 0.80. More importantly,
the package's requested futility rule uses only $P_{n_{\max},\ell}<F_\ell$;
ThermoCool AF required both predictive probabilities below 0.01. The package
also evaluates `Qn` at each scheduled look without a separate 4.5-month/50%
information gate.

## Source and status of modeled inputs

The labels in the following table are used consistently throughout the
vignette:

-   **reported**: stated numerically in a public primary source;
-   **inferred**: calculated from, or used to encode, reported information;
-   **assumed**: selected for this runnable package example;
-   **unavailable**: needed for exact reconstruction but not publicly supplied
    or not representable in the current package interface.

```{r input-audit, echo = FALSE}
input_audit <- data.frame(
  Input = c(
    "`N_total`",
    "`interim_look`",
    "`Qn`",
    "`Sn`",
    "`Fn`",
    "Futility statistic",
    "Early-claim information gate",
    "Non-stopping 106-patient look",
    "`prob_ha`",
    "`method`, `alternative`, `h0`",
    "`prior_bin`",
    "`cutpoints`, `generation_cutpoints`, `end_of_study`",
    "Sponsor predictive hazard prior",
    "`prior_surv`, `prior_surv_final`",
    "`block`, `rand_ratio`",
    "`lambda`, `lambda_time`",
    "Failure-time generator",
    "Worked benefit scenario",
    "Published OC scenario grid",
    "Observed enrollment history",
    "Observed interval hazards",
    "Complete regulatory sensitivity suite",
    "`prop_loss`",
    "`imputed_final`",
    "`N_impute`",
    "Evaluated `N_trials`"
  ),
  Value = c(
    "230",
    "150, 175, 200",
    "0.99 at each look",
    "0.90, 0.80, 0.80",
    "0.01 at each look",
    "Package uses P[nmax,l] only",
    "Not represented",
    "Not represented as an actionable look",
    "0.98",
    "bayes-bin, less, 0",
    "Beta(1, 1) in both arms",
    "0.5 and 2 months; 9-month horizon",
    "Three arm-specific piecewise rates with a hierarchical prior",
    "Fixed Gamma(1, 1) approximation in each interval and arm",
    "11; control 4 : treatment 7",
    "2, 3, 5 patients/month; changes at months 2 and 4",
    "Base hazards 0.65, 0.161, 0.05/month, scaled to target success",
    "Treatment 0.45; control 0.20 chronic success",
    "Three null and eight benefit scenarios",
    "11 after year 1; 53 after year 2; 160 at analysis; 167 at close",
    "June 2008 failures divided by exposure within each interval",
    "Not publicly available in full",
    "0 in each arm",
    "TRUE",
    "100 for one trial; 40 for small OC simulations; 5,000 in validation template",
    "50 per scenario"
  ),
  Status = c(
    "reported", "reported", "reported", "reported", "reported",
    "assumed", "reported", "reported", "reported", "inferred",
    "reported", "reported", "reported", "assumed", "reported",
    "reported", "reported", "reported", "reported", "reported",
    "inferred", "unavailable", "assumed", "inferred", "assumed",
    "assumed"
  ),
  `Source and mapping` = c(
    "JAMA Statistical Methods and FDA advisory briefing",
    "JAMA Statistical Methods and FDA advisory briefing",
    "Reported early-claim boundary; package applies its strict > rule",
    "Reported expected-success accrual boundaries",
    "Reported numeric futility boundary",
    "The requested enhancement differs from ThermoCool's P[n,l] and P[nmax,l] rule",
    "FDA briefing and transcript report 4.5 months or 50% endpoint-complete",
    "FDA requested it after the mid-trial amendment; zero probability of stopping",
    "Historical success was at least 0.98; the package classifies success using its strict > comparison",
    "Failure is the modeled binary event, so benefit is a lower treatment failure probability",
    "FDA advisory transcript and sponsor briefing",
    "JAMA and FDA describe breaks at 2 weeks/0.5 months and 2 months",
    "FDA briefing discloses Exp(rate = 1) priors on the Gamma hyperparameters, but not a directly reproducible fixed package prior",
    "Assumed plug-in approximation that sets both Gamma hyperparameters to the mean, 1, of their disclosed Exp(rate = 1) hyperpriors; not the sponsor's hierarchical prior",
    "JAMA Study Design",
    "Berry et al., Section 5.8, p. 245; default simulation accrual schedule",
    "Berry et al., Section 5.8, pp. 244--245; common shape scaled to an arm-specific nine-month success probability",
    "One of the benefit cases in Berry et al., Table 5.19; selected for the worked example",
    "Berry et al., Tables 5.18--5.19",
    "FDA sponsor briefing; realized history retained only for comparison",
    "Derived from FDA SSED Table 8; descriptive observed-data rates, not planning assumptions",
    "The FDA briefing redacts its scenario table, and Berry et al. notes additional sensitivity simulations without enumerating all of them",
    "No random loss generator is reported for the published scenarios; observed exclusions are not equivalent to random censoring",
    "JAMA reports multiple imputation for incomplete outcomes; the available package analysis is an approximation",
    "The evaluated counts are run-time choices; Berry et al. reports 1,000 burn-in and 5,000 retained MCMC iterations, which are not identical to the package computation",
    "Illustrative choice; the sponsor briefing reports 10,000 trials per scenario, while Berry et al. reports 25,000 for its tabulated null simulations"
  ),
  check.names = FALSE
)

knitr::kable(input_audit, format = "pipe")
```

## Mapping the longitudinal analysis model

The completed endpoint is binary, but its status is learned over time. The
trial predicted unknown nine-month outcomes with an arm-specific
piecewise-exponential time-to-failure model. The three intervals were
0--0.5 months, 0.5--2 months, and 2--9 months. After imputation, a patient's
failure time was reduced to a binary chronic-failure indicator, and the
completed-data analysis compared arm-specific success probabilities.

`goldilocks` represents this structure with `method = "bayes-bin"` and
`binary_imputation = "event-time"`. The package is parameterized on event
probability, so the code models chronic **failure** rather than chronic success:

$$
p_{\text{failure}} = 1 - p_{\text{chronic success}}.
$$

Consequently, superiority of ablation is encoded as

$$
\Pr(p_{T,\text{failure}} - p_{C,\text{failure}} < 0 \mid \mathcal D) > 0.98,
$$

using `alternative = "less"` and `h0 = 0`. Independent uniform priors on the
published chronic-success probabilities become the same independent
`Beta(1, 1)` priors after complementing to failure probabilities.

```{r analysis-model}
analysis_cutpoints_month <- c(0.5, 2)
effectiveness_horizon_month <- 9

prior_bin <- c(1, 1)

# Assumed plug-in approximation using the means of the disclosed Exp(1)
# hyperpriors; this is not the sponsor's hierarchical hazard prior.
prior_surv <- c(shape = 1, rate = 1)
```

The fixed `prior_surv` above approximates the sponsor's analysis prior. It is
separate from the true hazards used to generate virtual trials when studying
the design's operating characteristics.

## Published design-generating assumptions

Berry et al. (2010, pp. 244--245) gives the default piecewise-exponential
failure generator that was not enumerated in JAMA or the public FDA summaries.
Its base monthly hazards are 0.65, 0.161, and 0.05. Over the three analysis
intervals, these give nine-month chronic success of approximately 0.40:

$$
\exp\{-0.5(0.65)-1.5(0.161)-7(0.05)\} \approx 0.40.
$$

For a target chronic-success probability $p$, all three base hazards are
multiplied by the same factor:

$$
\theta_j(p) = \theta_j^*\frac{\log(p)}{\log(0.40)}.
$$

Thus, scenario probabilities change the overall failure level while preserving
the shape of the failure-time distribution.

```{r published-hazard-generator}
base_chronic_success <- 0.40
base_failure_hazard <- c(
  `0--0.5` = 0.65,
  `0.5--2` = 0.161,
  `2--9` = 0.05
)

hazard_for_success <- function(p) {
  stopifnot(length(p) == 1L, is.finite(p), p > 0, p < 1)
  base_failure_hazard * log(p) / log(base_chronic_success)
}

published_scenarios <- data.frame(
  Type = c(rep("Null", 3), rep("Benefit", 8)),
  `Treatment chronic success` = c(
    0.20, 0.40, 0.60,
    0.30, 0.40, 0.45, 0.50, 0.50, 0.60, 0.65, 0.70
  ),
  `Control chronic success` = c(
    0.20, 0.40, 0.60,
    0.20, 0.20, 0.20, 0.20, 0.40, 0.40, 0.40, 0.40
  ),
  check.names = FALSE
)

knitr::kable(published_scenarios, digits = 2)
```

Tables 5.18 and 5.19 of Berry et al. report the three null cases and eight
benefit cases shown above. There was therefore no single treatment-control
event-rate assumption. For the runnable example, we select the published
0.45-versus-0.20 chronic-success case, which the monograph discusses
specifically:

```{r worked-planning-scenario}
worked_chronic_success <- c(treatment = 0.45, control = 0.20)

hazard_treatment <- hazard_for_success(worked_chronic_success["treatment"])
hazard_control <- hazard_for_success(worked_chronic_success["control"])

planning_hazards <- data.frame(
  Arm = c("Catheter ablation", "ADT control"),
  `Target nine-month chronic success` = unname(worked_chronic_success),
  `Hazard 0--0.5 months` = c(hazard_treatment[1], hazard_control[1]),
  `Hazard 0.5--2 months` = c(hazard_treatment[2], hazard_control[2]),
  `Hazard 2--9 months` = c(hazard_treatment[3], hazard_control[3]),
  check.names = FALSE
)

knitr::kable(planning_hazards, digits = 3)
```

The same source reports a default simulation accrual rate of two patients per
month for the first two months, three per month for the next two months, and
five per month from the fifth month onward (Berry et al., 2010, p. 245).

```{r accrual-model}
enrollment_rate_per_month <- c(
  months_0_to_2 = 2,
  months_2_to_4 = 3,
  months_4_plus = 5
)
enrollment_rate_change_month <- c(2, 4)

data.frame(
  `Trial-calendar interval` = c("0--2 months", "2--4 months", "4+ months"),
  `Patients per month` = enrollment_rate_per_month,
  check.names = FALSE
)
```

These are reported defaults for simulations of the amended adaptive design.
`goldilocks` maps the three rates to its piecewise Poisson arrival process; the
monograph reports the rate schedule but does not specify enough operational
detail to verify that its enrollment-time generator was identical.
They should not be interpreted as the original 2004 fixed-design forecast: the
adaptive proposal was developed after accrual had begun and 106 patients were
already enrolled.

## Assumptions versus the reported trial history

The [FDA Summary of Safety and Effectiveness Data, Table 8, p. 13](https://www.accessdata.fda.gov/cdrh_docs/pdf3/P030031S011B.pdf)
reports June 2008 exposure and failure counts for the same three intervals.
Dividing failures by exposure gives observed-data rates, not prospective
design assumptions:

```{r descriptive-hazards}
fda_interval_data <- data.frame(
  arm = rep(c("treatment", "control"), each = 3),
  interval = rep(c("0--0.5", "0.5--2", "2--9"), 2),
  exposure_months = c(40.21, 104.17, 413.09, 23.27, 54.21, 90.46),
  failures = c(26, 3, 7, 13, 14, 20)
)
fda_interval_data$hazard_per_month <- with(
  fda_interval_data,
  failures / exposure_months
)

knitr::kable(fda_interval_data, digits = 4)

observed_hazard_treatment <- subset(
  fda_interval_data,
  arm == "treatment"
)$hazard_per_month
observed_hazard_control <- subset(
  fda_interval_data,
  arm == "control"
)$hazard_per_month
```

The planning and observed interval rates differ in shape as well as level:

```{r hazard-comparison}
hazard_comparison <- data.frame(
  Interval = names(base_failure_hazard),
  `Planning: treatment` = unname(hazard_treatment),
  `Observed: treatment` = observed_hazard_treatment,
  `Planning: control` = unname(hazard_control),
  `Observed: control` = observed_hazard_control,
  check.names = FALSE
)

knitr::kable(hazard_comparison, digits = 3)
```

Inserting the observed rates into a continuous piecewise-exponential
distribution gives a stronger treatment contrast than the selected planning
case. The JAMA Kaplan--Meier estimates were also more favorable to ablation:

```{r reported-probability-comparison}
interval_length_month <- diff(c(
  0,
  analysis_cutpoints_month,
  effectiveness_horizon_month
))

chronic_success_comparison <- data.frame(
  Arm = c("Catheter ablation", "ADT control"),
  `Published worked scenario` = unname(worked_chronic_success),
  `Implied by observed FDA interval rates` = c(
    exp(-sum(observed_hazard_treatment * interval_length_month)),
    exp(-sum(observed_hazard_control * interval_length_month))
  ),
  `JAMA Kaplan-Meier estimate` = c(0.66, 0.16),
  `FDA SSED Kaplan-Meier estimate` = c(0.64, 0.16),
  check.names = FALSE
)

knitr::kable(chronic_success_comparison, digits = 3)
```

The FDA-rate probabilities are descriptive approximations. They need not equal
the Kaplan--Meier estimates because they insert raw interval rates into a
continuous distribution without the sponsor's posterior calculation, handling
of time-zero failures, or analysis-population rules. The 0.66 and 0.64
ablation estimates are also source- and cutoff-specific rather than
interchangeable: JAMA reports 0.66, whereas the FDA SSED reports 0.64.

The published accrual model also differs sharply from the early operational
history reported in the
[FDA sponsor briefing, internal p. 128](https://web.archive.org/web/20170222051859/https://www.fda.gov/ohrms/dockets/ac/08/briefing/2008-4393b1-01-%20%20Sponsors%20Executive%20Summary.pdf).
Under the package convention that the first participant enrolls at time zero,
the default simulation model expects about 51 participants by month 12 and 111
by month 24. The trial had only 11 and 53, respectively, before enrollment
accelerated. It reached 160 at the September 2007 analysis and 167 when
enrollment closed the following month:

```{r accrual-comparison}
expected_enrollment <- function(month) {
  interval_time <- c(
    min(month, 2),
    max(min(month - 2, 2), 0),
    max(month - 4, 0)
  )
  1 + sum(enrollment_rate_per_month * interval_time)
}

accrual_comparison <- data.frame(
  Milestone = c(
    "After year 1",
    "After year 2",
    "First planned analysis",
    "Enrollment close"
  ),
  `Approximate trial month` = c(12, 24, 35, 36),
  `Reported cumulative enrollment` = c(11, 53, 160, 167),
  `Expected under published simulation default` = vapply(
    c(12, 24, 35, 36),
    expected_enrollment,
    numeric(1)
  ),
  check.names = FALSE
)

knitr::kable(accrual_comparison, digits = 0)
```

For an observed-history sensitivity analysis based on the reported year-end
totals, the corresponding piecewise average rates are:

```{r observed-accrual-sensitivity}
observed_enrollment_rate_per_month <- c(
  year_1 = (11 - 1) / 12,
  year_2 = (53 - 11) / 12,
  year_3 = (167 - 53) / 12
)
observed_enrollment_rate_change_month <- c(12, 24)

data.frame(
  `Trial-calendar interval` = c(
    "0--12 months",
    "12--24 months",
    "24--36 months"
  ),
  `Observed-history patients per month` =
    observed_enrollment_rate_per_month,
  check.names = FALSE
)
```

The analysis-month values are approximate calendar offsets from the first
enrollment in October 2004. Relative to the published 2, 3, and 5 per-month
default, the observed-history reconstruction is approximately 0.83, 3.50, and
9.50 per month: much slower during the first two years and substantially faster
late in the trial. The year-three rate uses the closing count of 167; 160 was
the preceding decision-analysis count rather than the final enrollment total.

## One simulated trial

The following analysis simulates one trial replicate under the published
0.45-versus-0.20 benefit scenario and default accrual schedule. It uses only
100 predictive imputations for illustration.
That is too few for regulatory calibration of boundaries as extreme as 0.99
and 0.01.

```{r one-trial}
set.seed(1030236)

thermocool_trial <- survival_adapt(
  hazard_treatment = hazard_treatment,
  hazard_control = hazard_control,
  cutpoints = analysis_cutpoints_month,
  generation_cutpoints = analysis_cutpoints_month,
  N_total = N_total,
  lambda = enrollment_rate_per_month,
  lambda_time = enrollment_rate_change_month,
  interim_look = interim_look,
  end_of_study = effectiveness_horizon_month,
  prior_surv = prior_surv,
  prior_surv_final = prior_surv,
  prior_bin = prior_bin,
  bin_method = "quadrature",
  binary_imputation = "event-time",
  block = 11,
  rand_ratio = c(control = 4, treatment = 7),
  prop_loss = 0,
  alternative = "less",
  h0 = 0,
  Fn = Fn,
  Sn = Sn,
  Qn = Qn,
  prob_ha = 0.98,
  N_impute = 100,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE,
  return_trace = TRUE
)

knitr::kable(
  thermocool_trial$summary[, c(
    "N_enrolled",
    "ppp_success",
    "stop_immediate_success",
    "stop_expected_success",
    "stop_futility",
    "trial_success",
    "stopping_reason",
    "decision_time"
  )],
  digits = 3,
  col.names = c(
    "Enrolled N", "Predictive success", "Immediate success stop",
    "Expected success stop", "Futility stop", "Trial success",
    "Stopping reason", "Decision time"
  )
)
```

`stop_immediate_success` is an official terminal success decision. It does not
wait for a later completed-follow-up analysis, so `post_prob_ha` and the final
effect estimate are `NA`. Binding futility is likewise an official terminal
failure. For compatibility, the package still attempts its historical
completed-follow-up diagnostic after futility; if that diagnostic cannot be
computed, `post_prob_ha` and `est_final` are `NA` and `trial_success` remains
`FALSE`.

The trace makes the ordered rule auditable:

```{r one-trial-trace}
knitr::kable(
  thermocool_trial$trace[, c(
    "look",
    "planned_N",
    "ppp_stop_now",
    "immediate_success_threshold",
    "success_threshold",
    "ppp_success_at_max",
    "futility_threshold",
    "decision"
  )],
  digits = 3,
  col.names = c(
    "Look", "Planned N", "PPSn", "Immediate success cut",
    "Expected success cut", "PPSmax", "Futility cut", "Decision"
  )
)
```

Here, PPSn is `ppp_stop_now`, PPSmax is `ppp_success_at_max`, and the three
cut columns show the immediate-success, expected-success, and futility thresholds.
Only looks actually reached appear in the trace. An immediate-success decision
prevents all subsequent looks.

## Small operating-characteristic demonstration

For simulation, collect common inputs in a named list and vary only the
data-generating hazards. The first scenario is the published 0.45-versus-0.20
benefit case. The second assigns the 0.20-success control profile to both arms,
which is one of the published null cases.

```{r small-oc}
thermocool_design <- list(
  cutpoints = analysis_cutpoints_month,
  generation_cutpoints = analysis_cutpoints_month,
  N_total = N_total,
  lambda = enrollment_rate_per_month,
  lambda_time = enrollment_rate_change_month,
  interim_look = interim_look,
  end_of_study = effectiveness_horizon_month,
  prior_surv = prior_surv,
  prior_surv_final = prior_surv,
  prior_bin = prior_bin,
  bin_method = "quadrature",
  binary_imputation = "event-time",
  block = 11,
  rand_ratio = c(control = 4, treatment = 7),
  prop_loss = 0,
  alternative = "less",
  h0 = 0,
  Fn = Fn,
  Sn = Sn,
  Qn = Qn,
  prob_ha = 0.98,
  N_impute = 40,
  empty_interval = "prior",
  method = "bayes-bin",
  imputed_final = TRUE,
  N_trials = 50,
  ncores = 2,
  return_trace = TRUE
)

thermocool_benefit <- do.call(sim_trials, c(
  thermocool_design,
  list(
    hazard_treatment = hazard_treatment,
    hazard_control = hazard_control,
    seed = 1030236
  )
))

thermocool_null <- do.call(sim_trials, c(
  thermocool_design,
  list(
    hazard_treatment = hazard_control,
    hazard_control = hazard_control,
    seed = 1030237
  )
))

oc_small <- summarise_sims(list(
  "Published benefit: 0.45 vs 0.20" = thermocool_benefit,
  "Published null: 0.20 vs 0.20" = thermocool_null
))

oc_display <- oc_small[, c(
  "scenario",
  "n_analyzed",
  "power",
  "stop_immediate_success",
  "stop_success",
  "stop_futility",
  "stop_max_N",
  "mean_N"
)]
names(oc_display)[names(oc_display) == "stop_success"] <-
  "stop_expected_success"

knitr::kable(
  oc_display,
  digits = 3,
  col.names = c(
    "Scenario", "Trials analyzed", "Power", "Immediate success stop",
    "Expected success stop", "Futility stop", "Maximum N", "Mean N"
  )
)
```

With only 50 trials and 40 imputations per look, these estimates have
substantial Monte Carlo error. They illustrate the mutually exclusive stopping
categories; they do not estimate the trial's reported operating
characteristics. `summarise_sims()` uses `stop_success` for stopping accrual for
expected success and reports immediate success separately as
`stop_immediate_success`.

The no-effect scenario provides a useful view of all four decision regions:

```{r stopping-plot, fig.width = 7, fig.height = 5}
plot_sim_stopping(thermocool_null)
```

```{r decision-plot, fig.width = 9, fig.height = 7}
plot_sim_decisions(thermocool_null)
```

In a decision plot, `Qn` adds a second horizontal boundary above `Sn`.
Observations above `Qn` are immediate successes. Observations between `Sn` and
`Qn` stop accrual for expected success. At or below `Sn`, `Fn` separates binding
futility from continued accrual.

## Operating-characteristic evaluation

A design that permits immediate declaration of success must be calibrated as a
whole. Neither `Qn = 0.99` nor `prob_ha = 0.98` automatically controls type I
error in a new design. Calibration must reflect all looks, the final analysis,
the binding futility rule, accrual, incomplete follow-up, the predictive model,
and Monte Carlo error.

The following unevaluated template increases both the number of simulated
trials and the predictive-imputation count. Its 5,000 predictive draws are the
closest package analogue to the 5,000 retained iterations described by Berry
et al.; the package does not reproduce the sponsor's hierarchical sampler or
its 1,000-iteration burn-in. The 25,000 trial replicates match the count stated
for Berry et al.'s null table; the sponsor briefing instead describes 10,000
trials per scenario, and the monograph does not state a replicate count for its
benefit table.

```{r full-validation, eval = FALSE}
thermocool_full_design <- modifyList(thermocool_design, list(
  N_trials = 25000,
  N_impute = 5000,
  ncores = 8,
  return_trace = FALSE
))

q_grid <- c(0.975, 0.99, 0.995, 1.00)

full_null <- lapply(seq_along(q_grid), function(i) {
  do.call(sim_trials, c(
    modifyList(thermocool_full_design, list(Qn = q_grid[i])),
    list(
      hazard_treatment = hazard_control,
      hazard_control = hazard_control,
      seed = 1031000 + i
    )
  ))
})
names(full_null) <- paste0("null 0.20 vs 0.20: Qn = ", q_grid)

full_benefit <- lapply(seq_along(q_grid), function(i) {
  do.call(sim_trials, c(
    modifyList(thermocool_full_design, list(Qn = q_grid[i])),
    list(
      hazard_treatment = hazard_treatment,
      hazard_control = hazard_control,
      seed = 1032000 + i
    )
  ))
})
names(full_benefit) <- paste0("benefit 0.45 vs 0.20: Qn = ", q_grid)

full_oc <- summarise_sims(c(full_null, full_benefit))
full_oc[, c(
  "scenario",
  "n_analyzed",
  "power",
  "power_mcse",
  "stop_immediate_success",
  "stop_immediate_success_mcse",
  "stop_success",
  "stop_futility",
  "stop_max_N",
  "mean_N",
  "mean_N_mcse"
)]
```

Useful additional scenarios include weaker treatment effects, different
control event rates, slower and faster accrual, arm-specific missingness,
alternative piecewise hazards, and violations of the predictive model. The
three published null cases and eight benefit cases in `published_scenarios`
provide a natural starting grid. The null design should still be evaluated over
a suitable nuisance-parameter range rather than at a single equal-arm profile.
Because a 0.99 boundary is estimated from imputations, the predictive draw
count should also be chosen so that decisions near `Qn`, `Sn`, and `Fn` are
sufficiently stable.

## What this example does and does not reproduce

The example preserves the central statistical idea:

-   one current-sample predictive probability drives two distinct upper
    decisions;
-   `Qn` declares success immediately, whereas `Sn` only stops accrual;
-   `Fn` acts on predictive success at the maximum sample size; and
-   all thresholds are evaluated in a prespecified order.

It uses the published default failure generator, accrual schedule, and one
published benefit scenario, but it does not reproduce the sponsor's complete
analysis, hierarchical hazard prior, confidential sensitivity suite,
treatment-specific evaluation-window origins, site-stratified randomization
sequences, protocol deviations, crossover, analysis-population exclusions,
operational overrun, or the original two-predictive-probability futility
requirement. It also does not recreate the mid-trial design amendment or its
statistical penalty. Those distinctions are why the simulated numerical
results should not be compared directly with the trial's regulatory analysis.

## Sources and document roles

-   Berry SM, Carlin BP, Lee JJ, Müller P.
    [*Bayesian Adaptive Methods for Clinical Trials*](https://doi.org/10.1201/EBK1439825488).
    Boca Raton, FL: Chapman & Hall/CRC; 2010. Section 5.8, "Case study:
    Ablation device to treat atrial fibrillation," pp. 241--246. This
    designer-authored case study supplies the default failure-time generator,
    its probability-scaling formula, the default accrual schedule, and the
    published operating-characteristic scenario grid used in this vignette.

-   Wilber DJ, Pappone C, Neuzil P, et al. [Comparison of antiarrhythmic drug
    therapy and radiofrequency catheter ablation in patients with paroxysmal
    atrial fibrillation: a randomized controlled trial](https://pubmed.ncbi.nlm.nih.gov/20103757/).
    *JAMA*. 2010;303(4):333-340. doi:10.1001/jama.2009.2029.
    The article reports the analysis boundaries and observed Kaplan--Meier
    results, but not the numerical simulation generator or accrual schedule.
-   U.S. Food and Drug Administration. [Summary of Safety and Effectiveness
    Data, PMA P030031/S011](https://www.accessdata.fda.gov/cdrh_docs/pdf3/P030031S011B.pdf).
    Table 8, p. 13 reports the June 2008 interval exposure and failures used
    above as observed-data comparisons, not planning assumptions.
-   U.S. Food and Drug Administration. [Circulatory System Devices Panel,
    November 20, 2008 meeting materials](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfAdvisory/details.cfm?mtg=705).
    This is the archive landing page for the FDA and sponsor summaries, slides,
    panel questions, and transcript.
-   U.S. Food and Drug Administration.
    [FDA Executive Summary, pp. 10--12](https://web.archive.org/web/20170222060225/https://www.fda.gov/ohrms/dockets/ac/08/briefing/2008-4393b1-01%20%20FDA%20executive%20summary%20FINAL.pdf).
    It documents the amendment, 106-patient non-stopping look, information
    gate, stopping rules, and hierarchical longitudinal model.
-   Biosense Webster, Inc.
    [Sponsor Executive Summary](https://web.archive.org/web/20170222051859/https://www.fda.gov/ohrms/dockets/ac/08/briefing/2008-4393b1-01-%20%20Sponsors%20Executive%20Summary.pdf).
    Internal pp. 24--25 and 128 report the simulation exercise and realized
    enrollment history. The briefing says that 10,000 trials per scenario were
    simulated, but its operating-characteristic table is redacted. Berry et al.
    later reports 25,000 simulations specifically for its tabulated null cases.
-   U.S. Food and Drug Administration. [PMA supplement P030031/S011 approval
    record](https://www.accessdata.fda.gov/scripts/cdrh/cfdocs/cfpma/pma.cfm?id=P030031S011).
    FDA issued the decision on February 6, 2009 for the expanded paroxysmal
    atrial-fibrillation indication.
