---
title: "Two-arm randomized trials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-arm randomized trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
load("vignette-sims.rda")
```

Broglio et al. (2014) presented a hypothetical trial example. We use a similar
setting and show how each statistical and operational assumption is represented
in `goldilocks`.

## Design assumptions

The setting is a two-arm trial with equal randomization to control or treatment.
The primary endpoint is overall survival (OS), measured from enrollment to
death from any cause or last follow-up. The simulation treats enrollment and
randomization as occurring at the same time. The assumed 12-month OS
probability in the control arm is 30%. The minimum and maximum sample sizes are
100 and 300, respectively, and no loss to follow-up is assumed. Each
participant is followed until death or 12 months, whichever comes first.
After an expected-success stop or enrollment of 300 participants, the primary
analysis is conducted once all enrolled participants' required event or
censoring outcomes are available. Its calendar time therefore depends on the
last observed event or censoring time.

From this information, we have:

-   Equal randomization: `block = 2` and `rand_ratio = c(control = 1, treatment = 1)` (default parameters)
-   Primary endpoint is at 12 months: `end_of_study = 12`
-   12-month event rate for control arm: `hazard_control = prop_to_haz(1 - 0.30, endtime = 12)` (note that the input argument is the failure proportion, not the survival proportion)
-   No change points in hazard: `cutpoints = NULL` (default parameter)
-   Maximum sample size: `N_total = 300`
-   No attrition: `prop_loss = 0`

Named arms make unequal allocation unambiguous. For example,
`rand_ratio = c(control = 1, treatment = 2)` specifies 1:2 randomization.
Unnamed values are interpreted in control-treatment order for backward
compatibility, but explicit arm names are recommended in a protocol simulation.

Sample-size selection analyses are planned after 100 participants have enrolled
and after each additional 25 participants. Futility stopping is allowed from
the first analysis, with $F_n=0.10$. Stopping accrual for expected success is
allowed from 125 participants onward, with $S_n=0.90$. The assumed enrollment
rate is constant at five participants per month.

Enrollment is stochastic even though the rate is constant. The package fixes
the first patient at calendar time zero and generates each later inter-arrival
gap from an exponential distribution with rate 5 per month. Consequently, the
expected time from the first to the 300th enrollment is $(300 - 1) / 5 = 59.8$
months, but the realized completion time differs between simulated trials.
`lambda_time = NULL` indicates that there are no enrollment-rate
changes; zero is implicit and must not be supplied.

For comparison, a ramp-up specification such as `lambda = c(2, 5)` and
`lambda_time = 6` assigns positive realized enrollment times in $(0,6]$ to 2
expected enrollments per month and later times to 5 per month. The first participant
at zero is the fixed calendar origin. Fractional changes such as
`lambda_time = 6.5` are also simulated exactly. Enrollment-rate knots use the
trial calendar measured from first participant in, whereas hazard `cutpoints` use
each participant's follow-up time measured from that participant's enrollment. The two
schedules are independent and need not share their knots.

From this information, we have:

-   Interim sample size looks: `interim_look = seq(100, 275, 25)`
-   Futility probability thresholds: `Fn = rep(0.10, 8)`
-   Predicted success probability thresholds: `Sn = c(1, rep(0.9, 7))`
-   Immediate success is disabled: `Qn = 1` (the default)
-   `lambda = 5` and `lambda_time = NULL` (default parameter)

Note that the first value of `Sn` is 1. This is because the trial is not allowed to stop for predicted success at the first interim analysis of $n = 100$. The remaining elements of `Sn` are 0.9, corresponding to 90%.

## Final analysis and hypotheses

The primary analysis is a two-sided log-rank test, with success declared at the $\alpha = 0.05$ level.

From this information, we have:

-   Two-sided log-rank test used: `alternative = "two.sided"` and `method = "logrank"`
-   $\alpha = 0.05$ level used to declare success: `prob_ha = 0.95`

For a frequentist analysis, `goldilocks` expresses evidence as $1-p$, so
`prob_ha = 0.95` corresponds to a two-sided significance level of 0.05. A
Bayesian analysis instead compares a posterior probability with `prob_ha`;
the common numerical scale does not make the frequentist and Bayesian decision
rules inferentially equivalent. The log-rank analysis requires `h0 = 0`,
corresponding to equality of the survival distributions.

### One-sided tests

The example above uses a two-sided test. A design targeting benefit in one
direction can instead prespecify a one-sided test. The `cox` and `logrank`
methods support all three alternatives via the `alternative` argument. For
these methods, the direction of benefit is:

-   `alternative = "less"` declares success when the treatment arm has a *lower* hazard (longer survival) than control.
-   `alternative = "greater"` declares success when the treatment arm has a *higher* hazard.

For instance, to run the same design as a one-sided log-rank test at the 0.025 level, we would set:

```{r one_sided_example, eval=FALSE}
out_power_1sided <- update(
  out_power,
  alternative = "less",
  prob_ha = 0.975
)
```

The frequentist binary risk-difference analyses support all three alternatives and compare $p_{\text{treatment}} - p_{\text{control}}$ with `h0`. Use `method = "riskdiff-fm"` for a Farrington-Manning score test that remains defined for sparse boundary tables, or `method = "riskdiff-wald"` for the plug-in Wald test. The Bayesian test (`method = "bayes-surv"`) requires a one-sided alternative (`"less"` or `"greater"`), and `"two.sided"` raises an error. For the Bayesian test the effect is measured on the cumulative-failure-probability scale, $p_{\text{treatment}} - p_{\text{control}}$ at `end_of_study`, compared against the margin `h0` (default `0`):

-   `alternative = "less"` declares success when the posterior probability that $p_{\text{treatment}} - p_{\text{control}} < h_0$ exceeds the threshold `prob_ha` -- i.e. the treatment arm has a failure probability lower than the `h0` margin relative to control. With the default `h0 = 0`, this means lower failure probability (longer survival) than control.
-   `alternative = "greater"` declares success when the posterior probability that $p_{\text{treatment}} - p_{\text{control}} > h_0$ exceeds `prob_ha`.

For `method = "rmst"`, the effect is instead treatment-minus-control restricted
mean survival time through a fixed `rmst_tau`. Longer survival corresponds to
`alternative = "greater"`. With time measured in months, `h0` is a difference
in months: use `h0 = 0` for superiority, or `h0 = -1` for non-inferiority
allowing a loss of one month of RMST. Choose the method, effect scale, direction,
and horizon together before evaluating the design. The
[RMST vignette](rmst.html) gives a worked example with a delayed treatment
effect and explains the support required through `rmst_tau`.

## Simulation settings

The operating characteristics will be determined using 500 simulated trials. At each interim analysis, we will use 100 imputations and assume independent weakly-informative $\operatorname{Gamma}(0.1, 0.1)$ prior distributions for the treatment and control arm event time hazard rate parameters. As this is computationally expensive overall, we will exploit the option to parallelize the simulations over multiple cores.

-   Number of simulated trials: `N_trials = 500`
-   Number of imputations from predictive distribution: `N_impute = 100`
-   Independent prior distribution for each hazard rate parameter: `prior_surv = c(0.1, 0.1)`
-   Parallel computation using eight cores: `ncores = 8`
-   Reproducible Monte Carlo study: `seed = 123`

The parameter `N_mcmc` is not used by the log-rank test. Here `prop_loss = 0`
means no dropout. A positive value would specify the CDF of an independent
exponential dropout time at `end_of_study`; actual censoring by dropout can be
less frequent because events can occur first. Log-rank, Cox, and RMST analyses retain
right-censored follow-up with `imputed_final = FALSE`, including when dropout
occurs. Imputed final analyses are not available for `method = "logrank"`.

For methods accepting `imputed_final = TRUE`, complete final outcomes use the
selected test directly. With missing outcomes, `"cox"`, `"rmst"`, and `"riskdiff-wald"`
support final imputation and Rubin pooling, requiring at least two imputations
and positive total variance. FM final imputation is unsupported; simulations
with `method = "riskdiff-fm"` and `imputed_final = TRUE` require zero dropout
in both arms. Binary
analyses with `imputed_final = FALSE` exclude incomplete endpoint statuses;
that complete-case analysis can be biased even under independent dropout,
because early events can be observed before dropout. Binary designs with
dropout should assess final imputation and its model assumptions.

## Power and type I error

Initially, we want to determine the power to detect a significant treatment effect when the OS rate at 12-months for the treatment arm is 50%.

```{r setup}
library(goldilocks)
```

```{r example_power, eval=FALSE}
hc <- prop_to_haz(0.7, endtime = 12)
ht <- prop_to_haz(0.5, endtime = 12)

out_power <- sim_trials(
  hazard_treatment = ht,
  hazard_control = hc,
  cutpoints = NULL,
  N_total = 300,
  lambda = 5,
  lambda_time = NULL,
  interim_look = seq(100, 275, 25),
  end_of_study = 12,
  prior_surv = c(0.1, 0.1),
  block = 2,
  rand_ratio = c(control = 1, treatment = 1),
  prop_loss = 0,
  alternative = "two.sided",
  Fn = rep(0.10, 8),
  Sn = c(1, rep(0.9, 7)),
  prob_ha = 0.95,
  N_impute = 100,
  N_trials = 500,
  method = "logrank",
  ncores = 8,
  seed = 123)
```

The 500 replicates used here are sufficient for illustration but not for a
definitive design decision. A larger simulation should be used when greater
precision is needed for type I error, power, or expected sample size.

To estimate type I error, we simulate under the null by setting the treatment
hazard equal to the control hazard. `update()` retains the remaining design
specification:

```{r example_type1, eval=FALSE}
out_t1error <- update(out_power, hazard_treatment = hc, seed = 124)
```

```{r summarise_sims}
initial_oc <- summarise_sims(list(out_power, out_t1error))
knitr::kable(
  initial_oc[c(
    "scenario",
    "n_requested",
    "n_used",
    "n_failed",
    "power",
    "stop_success",
    "stop_futility",
    "stop_max_N",
    "mean_N"
  )],
  digits = 3,
  col.names = c(
    "Scenario", "Requested", "Used", "Failed runs", "Power",
    "Expected success stop", "Futility stop", "Maximum N", "Mean N"
  ),
  caption = "Operating characteristics with a two-sided log-rank test at the 0.05 level. Scenario 1 is the alternative (treatment OS 50%); scenario 2 is the null (treatment OS 30%)."
)
```

The estimated type I error under this design is the `power` value for scenario
2: `r sprintf("%.1f%%", 100 * initial_oc$power[2])`. Its 95% Wilson Monte Carlo
interval is `r sprintf("%.1f%% to %.1f%%", 100 * initial_oc$power_mc_lower[2], 100 * initial_oc$power_mc_upper[2])`.
The point estimate alone does not establish whether the design exceeds the
intended 0.05 level; both Monte Carlo uncertainty and the complete adaptive
decision rule matter.

## Calibrating the final threshold

The final-analysis threshold should therefore be calibrated jointly with the
interim rules. As a preliminary candidate, consider $P < 0.04$, specified as
`prob_ha = 0.96`. The [calibration vignette](calibrating-prob-ha.html) gives a
systematic grid-search and independent-validation procedure. The candidate
below illustrates a stricter threshold; it is not a validated calibration.

```{r example_p0.04, eval=FALSE}
out_power2 <- update(out_power, prob_ha = 0.96, return_trace = TRUE)
out_t1error2 <- update(
  out_power2,
  hazard_treatment = hc,
  return_trace = FALSE,
  seed = 125
)
```

```{r summarise_sims_p0.04}
oc_calibrated <- summarise_sims(list(
  "target: treatment OS 50%" = out_power2,
  "null: treatment OS 30%" = out_t1error2
), max_mcse = c(power = 0.02, mean_N = 3))

target_oc <- oc_calibrated[
  oc_calibrated$scenario == "target: treatment OS 50%",
]
null_oc <- oc_calibrated[
  oc_calibrated$scenario == "null: treatment OS 30%",
]

format_mc_interval <- function(estimate, lower, upper, digits = 3) {
  format_string <- paste0(
    "%.", digits, "f [%.", digits, "f-%.", digits, "f]"
  )
  sprintf(format_string, estimate, lower, upper)
}
oc_calibrated_display <- data.frame(
  scenario = oc_calibrated$scenario,
  simulations = sprintf(
    "%d/%d (%d)",
    oc_calibrated$n_used,
    oc_calibrated$n_requested,
    oc_calibrated$n_failed
  ),
  power = format_mc_interval(
    oc_calibrated$power,
    oc_calibrated$power_mc_lower,
    oc_calibrated$power_mc_upper
  ),
  expected_success = format_mc_interval(
    oc_calibrated$stop_success,
    oc_calibrated$stop_success_mc_lower,
    oc_calibrated$stop_success_mc_upper
  ),
  futility = format_mc_interval(
    oc_calibrated$stop_futility,
    oc_calibrated$stop_futility_mc_lower,
    oc_calibrated$stop_futility_mc_upper
  ),
  maximum_N = format_mc_interval(
    oc_calibrated$stop_max_N,
    oc_calibrated$stop_max_N_mc_lower,
    oc_calibrated$stop_max_N_mc_upper
  ),
  mean_N = format_mc_interval(
    oc_calibrated$mean_N,
    oc_calibrated$mean_N_mc_lower,
    oc_calibrated$mean_N_mc_upper,
    digits = 1
  )
)
knitr::kable(
  oc_calibrated_display,
  col.names = c(
    "Scenario",
    "Used/requested (failed)",
    "Power [95% MC CI]",
    "Expected success [95% MC CI]",
    "Futility [95% MC CI]",
    "Maximum N [95% MC CI]",
    "Mean N [95% MC CI]"
  ),
  caption = "Operating characteristics with the more stringent P < 0.04 threshold (`prob_ha = 0.96`)."
)
```

Here, "95% MC CI" means a Monte Carlo confidence interval: it describes how
precisely this finite batch estimates the operating characteristic under the
fixed simulation assumptions. It is **not a clinical confidence interval for
the treatment effect** and does not represent uncertainty in the assumed event,
accrual, or loss-to-follow-up models. Probability intervals use the Wilson
method, while mean sample size uses a t interval based on its Monte Carlo
standard error. The optional `max_mcse` argument warns when a named precision
target is not met; it does not change the simulations or estimates.

In this illustrative `r target_oc$n_used`-trial simulation, assuming a
50% 12-month OS probability in the treatment arm,
`r sprintf("%.1f%%", 100 * target_oc$stop_success)` of trials stopped
accrual for expected success,
`r sprintf("%.1f%%", 100 * target_oc$stop_futility)` stopped for futility,
and the mean sample size was `r sprintf("%.1f", target_oc$mean_N)`.
Estimated power was `r sprintf("%.1f%%", 100 * target_oc$power)`.
Under the null scenario, in which treatment and control had the same 12-month
OS probability, `r sprintf("%.1f%%", 100 * null_oc$stop_futility)`
stopped for futility. Larger simulation
studies are appropriate when the displayed Monte Carlo precision is
insufficient for a final design decision.

## Calendar time and follow-up

The same simulation can be summarized on the calendar-time scale without
adding any design arguments. Time zero is first patient enrolled, and the time
unit is months in this example. "Analysis ready" is when the last observed
event or censoring required for the final analysis becomes available; it does
not include an external allowance for data cleaning or database lock. The
percentage in the trials column uses all requested simulations as its
denominator, so failed and excluded simulations cannot silently disappear.

```{r calendar-duration}
calendar_oc <- summarise_calendar_time(out_power2)
calendar_duration <- calendar_oc$trial_duration
calendar_duration$trials <- sprintf(
  "%d (%.1f%%)",
  calendar_duration$n_trials,
  calendar_duration$percent_trials
)
calendar_duration$accrual <- sprintf(
  "%.1f [%.1f-%.1f]",
  calendar_duration$accrual_stop_median,
  calendar_duration$accrual_stop_p10,
  calendar_duration$accrual_stop_p90
)
calendar_duration$analysis_ready <- sprintf(
  "%.1f [%.1f-%.1f]",
  calendar_duration$analysis_ready_median,
  calendar_duration$analysis_ready_p10,
  calendar_duration$analysis_ready_p90
)
knitr::kable(
  calendar_duration[c(
    "stopping_reason",
    "trials",
    "mean_N",
    "accrual",
    "analysis_ready",
    "followup_person_time_mean",
    "peak_active_followup_mean"
  )],
  digits = 1,
  col.names = c(
    "Stopping reason",
    "Trials, n (%)",
    "Mean enrolled",
    "Accrual stopped, median [P10-P90]",
    "Analysis ready, median [P10-P90]",
    "Mean person-months",
    "Mean peak under follow-up"
  ),
  caption = "Calendar-time duration and follow-up burden under the treatment-effect scenario."
)
```

Because `out_power2` was simulated with `return_trace = TRUE`, a second wide
table describes when each interim look was reached and how many subjects were
actively under follow-up at that time. A trial that stops before a later look
remains in the requested denominator but does not contribute a timing value at
that look.

```{r calendar-interims}
calendar_interim <- calendar_oc$interim_timing
calendar_interim$reached <- sprintf(
  "%d (%.1f%%)",
  calendar_interim$n_reached,
  calendar_interim$percent_reached
)
calendar_interim$calendar_time <- sprintf(
  "%.1f [%.1f-%.1f]",
  calendar_interim$calendar_time_median,
  calendar_interim$calendar_time_p10,
  calendar_interim$calendar_time_p90
)
calendar_interim$active_followup <- sprintf(
  "%.0f [%.0f-%.0f]",
  calendar_interim$active_followup_median,
  calendar_interim$active_followup_p10,
  calendar_interim$active_followup_p90
)
knitr::kable(
  calendar_interim[c(
    "look",
    "planned_N",
    "reached",
    "calendar_time",
    "active_followup"
  )],
  col.names = c(
    "Look",
    "Planned N",
    "Reached, n (%)",
    "Calendar month, median [P10-P90]",
    "Active follow-up, median [P10-P90]"
  ),
  caption = "Calendar timing and concurrent follow-up at each interim look."
)
```

## Graphical assessment

The same results can be viewed graphically. `plot_sim_ocs()` compares final
success, stopping behavior, and expected sample size across the treatment-effect
scenarios. Because the meaning and direction of an effect depends on the chosen
analysis, the effect scale is supplied explicitly; here it is the true 12-month
treatment survival probability. Match values by scenario name so that they
remain aligned when `summarise_sims()` sorts the rows.

```{r plot-ocs, fig.width=9, fig.height=4.5}
effect_by_scenario <- c(
  "target: treatment OS 50%" = 0.50,
  "null: treatment OS 30%" = 0.30
)
oc_calibrated$true_treatment_survival <- unname(
  effect_by_scenario[oc_calibrated$scenario]
)
plot_sim_ocs(
  oc_calibrated,
  effect = "true_treatment_survival",
  xlab = "True 12-month treatment survival probability"
)
```

For a single scenario, `plot_sim_stopping()` can show four complementary
views. The default marginal view gives each outcome as a percentage of all
simulated trials. The conditional view uses only trials still active when each
look begins as its denominator, while the cumulative view shows the status of
all trials after every look and includes those continuing to the next look. A
flowchart view displays counts moving from the total simulation set
through futility, continued enrollment, and early success at successive looks.
Because `out_power2` retains simulation traces, the latter three views include
reached looks even when no trial stopped at that look.
Percentage labels use a compact size so values at adjacent looks remain
visually distinct.

```{r plot-stopping, fig.width=8, fig.height=5.5, out.width='100%'}
plot_sim_stopping(out_power2)
```

```{r plot-stopping-conditional, fig.width=8, fig.height=5.5, out.width='100%'}
plot_sim_stopping(out_power2, type = "conditional")
```

```{r plot-stopping-cumulative, fig.width=8, fig.height=5.5, out.width='100%'}
plot_sim_stopping(out_power2, type = "cumulative")
```

```{r plot-stopping-flowchart, fig.width=7, fig.height=12, out.width='100%'}
plot_sim_stopping(out_power2, type = "flowchart")
```

The predictive-probability decision map requires traces from every simulated
trial. These are opt-in because they increase the size of the simulation
result:

```{r plot-decisions, eval=FALSE}
plot_sim_decisions(out_power2)
```

Each decision-map panel represents an interim look. The horizontal coordinate
is the predictive probability of success after continuing to the maximum sample
size; the vertical coordinate is the predictive probability if enrollment
stops now. Shading and dashed lines show the continuation, futility, and
expected-success regions.

Once we have identified a suitable design, we would typically re-run the simulations using a larger number of simulations and, perhaps, imputations.

## References

Broglio KR, Connor JT, Berry SM. Not too big, not too small: a Goldilocks approach to sample size selection. *Journal of Biopharmaceutical Statistics*, 2014; **24(3)**: 685–705.
